---
title: "Visualizing a Single Quantitative Variable"
subtitle: "TEMBA Statistics 2026"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")

austin_full <- load_austin_full()
austin <- load_austin_subset()
ercot <- load_ercot_daily()
returns <- load_returns()

austin_zip <- austin_full |>
  filter(zipcode == 78752)
```

## A histogram counts observations in bins

```{r}
#| label: viz-hist-price-count
#| fig-width: 9
#| fig-height: 4.4
n_bins <- 20
breaks <- seq(min(austin_zip$latest_price), max(austin_zip$latest_price),
              length.out = n_bins + 1)
bw <- diff(breaks)[1]

hist_df <- austin_zip |>
  mutate(bin_idx = pmin(findInterval(latest_price, breaks, rightmost.closed = TRUE),
                        n_bins)) |>
  count(bin_idx) |>
  mutate(
    lower  = breaks[bin_idx],
    upper  = breaks[bin_idx + 1],
    mid    = (lower + upper) / 2,
    tip    = paste0(dollar(round(lower), accuracy = 1000), "–",
                   dollar(round(upper), accuracy = 1000), "\n",
                   n, " home", ifelse(n == 1, "", "s"))
  )

ggplot(hist_df, aes(x = mid, y = n, width = bw)) +
  geom_col(fill = "steelblue", color = "white") +
  geom_rug(data = austin_zip, aes(x = latest_price), inherit.aes = FALSE,
           alpha = 0.4) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Count")
```

`r nrow(austin_zip)` home sales in Austin ZIP 78752. Each tick on the baseline is one sale.

## The picture depends on the bin width

```{r}
#| label: viz-binwidth
#| fig-width: 10
#| fig-height: 3.6
mk_hist <- function(bins = NULL, binwidth = NULL, boundary = NULL, title = "") {
  ggplot(austin_zip, aes(x = latest_price)) +
    geom_histogram(bins = bins, binwidth = binwidth, boundary = boundary,
                   fill = "steelblue", color = "white") +
    geom_rug(alpha = 0.3) +
    scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(x = NULL, y = NULL, title = title) +
    theme_minimal(base_size = 12)
}

mk_hist(bins = 10, title = "10 bins") |
  mk_hist(bins = 25, title = "25 bins") |
  mk_hist(bins = 60, title = "60 bins")
```

Same `r nrow(austin_zip)` sales in every panel.

## A density plot is a smoothed histogram

```{r}
#| label: viz-density-price
#| fig-width: 9
#| fig-height: 4.4
ggplot(austin_zip, aes(x = latest_price)) +
  geom_density(fill = "steelblue", alpha = 0.3) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Density") +
  geom_rug(alpha = 0.4)
```

The curve is high where there are more sales close to that price.

## Areas under the curve are proportions of the data

```{r}
#| label: viz-density-area
#| fig-width: 9
#| fig-height: 4.4
load_vals <- ercot$load
dens <- density(load_vals)
dens_df <- tibble(x = dens$x, y = dens$y)

kde_area <- function(lo, hi) {
  seg <- filter(dens_df, x >= lo, x <= hi)
  sum(diff(seg$x) * (head(seg$y, -1) + tail(seg$y, -1)) / 2)
}

q_mid  <- quantile(load_vals, c(0.40, 0.50))
q_tail <- quantile(load_vals, c(0.90, 1.00))
area_mid  <- kde_area(q_mid[1], q_mid[2])
area_tail <- kde_area(q_tail[1], q_tail[2])

ggplot(dens_df, aes(x = x, y = y)) +
  geom_area(data = filter(dens_df, x >= q_mid[1], x <= q_mid[2]),
            fill = "darkorange", alpha = 0.8) +
  geom_area(data = filter(dens_df, x >= q_tail[1], x <= q_tail[2]),
            fill = "darkorange", alpha = 0.8) +
  geom_line(color = "steelblue", linewidth = 1) +
  annotate("text", x = mean(q_mid), y = max(dens_df$y) * 1.08,
           label = paste0("area = ", number(area_mid, accuracy = 0.001),
                          "\n10% of days"),
           size = 3.4, lineheight = 1.1) +
  annotate("segment", x = mean(q_mid), xend = mean(q_mid),
           y = max(dens_df$y) * 1.0, yend = max(dens_df$y) * 0.9) +
  annotate("text", x = mean(q_tail), y = max(dens_df$y) * 0.32,
           label = paste0("area = ", number(area_tail, accuracy = 0.001),
                          "\n10% of days"),
           size = 3.4, lineheight = 1.1) +
  scale_x_continuous(labels = comma_format()) +
  labs(x = "Daily mean load (MW)", y = "Density")
```

ERCOT DFW daily mean load, 2016--2020 (`r comma(nrow(ercot))` days): the upper interval is over four times wider for the same 10% share ([source](https://github.com/blakeshaffer/ercotproject)).

## Densities have a bandwidth instead of bins

```{r}
#| label: viz-density-bw
#| fig-width: 10
#| fig-height: 3.6
mk_dens <- function(adjust, title) {
  ggplot(austin_zip, aes(x = latest_price)) +
    geom_density(adjust = adjust, fill = "steelblue", alpha = 0.3) +
    geom_rug(alpha = 0.3) +
    scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(x = NULL, y = NULL, title = title) +
    theme_minimal(base_size = 12)
}

mk_dens(0.3, "Wiggly (adjust = 0.3)") |
  mk_dens(1,   "Default (adjust = 1)") |
  mk_dens(2.5, "Smooth (adjust = 2.5)")
```

A single interpretable knob, with a default computed from the data rather than picked by hand.

## Raw counts can't compare groups of different sizes

```{r}
#| label: viz-hist-bigsmall
#| fig-width: 9
#| fig-height: 4
austin_full |>
  filter(zipcode %in% c(78748, 78722), latest_price < 1050000) |>
  mutate(zipcode = factor(zipcode,
                          levels = c(78748, 78722),
                          labels = c("78748 (1,100 sales)", "78722 (90 sales)"))) |>
  ggplot(aes(x = latest_price)) +
  geom_histogram(binwidth = 50000, boundary = 0,
                 fill = "steelblue", color = "white") +
  facet_wrap(~zipcode) +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Sale price", y = "Count")
```

Same \$50,000 bins and the same count axis, but 78748 has more than ten times as many sales as 78722, so the smaller panel is squashed flat.

## Rescaled to density, the shapes are comparable

```{r}
#| label: viz-hist-density-overlay
#| fig-width: 9
#| fig-height: 4
austin_full |>
  filter(zipcode %in% c(78748, 78722), latest_price < 1050000) |>
  mutate(zipcode = factor(zipcode,
                          levels = c(78748, 78722),
                          labels = c("78748 (1,100 sales)", "78722 (90 sales)"))) |>
  ggplot(aes(x = latest_price)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 50000, boundary = 0,
                 fill = "steelblue", color = "white") +
  geom_density(color = "darkorange", linewidth = 1) +
  facet_wrap(~zipcode) +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Sale price", y = "Density")
```

Both panels share one vertical scale: prices in 78722 sit higher and spread wider than in 78748.

## Percentiles

::: {.incremental}
- Suppose you scored 655 on the GMAT Focus Edition.
- Your score report says you're at the 90th percentile.
- That means you did at least as well as about 90% of test takers: 655 is the **90th percentile** of GMAT scores.
- In general, the **p-th percentile** is the value at or below which p% of the observations fall.
:::

## Median, quartiles, quantiles

::: {.incremental}
- The **median** is the 50th percentile: half the observations fall below it.
- The **quartiles** Q1, Q2, Q3 are the 25th, 50th, and 75th percentiles --- they split the data into four equal parts.
- Written as a proportion instead of a percent, a percentile is a **quantile**: the 90th percentile is the 0.90 quantile, and the median is the 0.50 quantile.
:::

## The empirical CDF plots every percentile at once

```{r}
#| label: viz-gmat-ecdf
#| fig-width: 9
#| fig-height: 4.4
ggplot(gmat, aes(x = score, y = quantile)) +
  geom_step(direction = "hv", color = "steelblue", linewidth = 1) +
  geom_point(size = 1.4, color = "steelblue") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "GMAT Focus Edition total score", y = "Cumulative percentage")
```

GMAC's full concordance, 2020--2025, is one curve: the fraction of test takers at or below each score.

## The ECDF of our sale prices

```{r}
#| label: viz-ecdf-price
#| fig-width: 9
#| fig-height: 4.4
ecdf_df <- austin_zip |>
  count(latest_price, name = "n_price") |>
  arrange(latest_price) |>
  mutate(ecdf = cumsum(n_price) / sum(n_price))

ggplot(ecdf_df, aes(x = latest_price, y = ecdf)) +
  geom_step(aes(group = 1), direction = "hv", color = "steelblue", linewidth = 1) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Sale price", y = "Cumulative percentage")
```

The height at each price is the fraction of sales at or below it. There are no bins or bandwidths to choose, and the quartiles read off the plot directly.

# Describing distributions

## Four things to look for

::: {.incremental}
- **Location** --- where is the center; what is a typical value?
- **Spread** --- how far do values range from the center?
- **Skewness** --- is the distribution symmetric, or stretched to one side?
- **Modality** --- how many peaks are there?
- And **outliers** --- observations far from the bulk of the data
:::

## Symmetric

```{r}
#| label: viz-returns-symmetric
#| fig-width: 8.5
#| fig-height: 4.4
max_abs <- max(abs(returns$market))
ggplot(returns, aes(x = market)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, fill = "steelblue", color = "white") +
  geom_density(alpha = 0.3) +
  geom_rug(alpha = 0.4) +
  scale_x_continuous(labels = percent_format(),
                     limits = c(-max_abs, max_abs)) +
  labs(x = "Weekly return", y = "Density")
```

Weekly S&P 500 returns, October 2023 to September 2025: the left and right halves are rough mirror images.

## Right-skewed

```{r}
#| label: viz-ercot-skewed
#| fig-width: 8.5
#| fig-height: 4.4
ggplot(ercot, aes(x = load)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40, fill = "steelblue", color = "white") +
  geom_density(alpha = 0.3) +
  scale_x_continuous(labels = comma_format()) +
  labs(x = "Load (MW)", y = "Density")
```

Daily mean electricity demand. The high-demand days in the right tail pull the mean above the median.

## Bimodal

```{r}
#| label: viz-austin-bimodal
#| fig-width: 8.5
#| fig-height: 4.4
ggplot(austin, aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "steelblue", color = "white") +
  geom_density(alpha = 0.3) +
  geom_rug(alpha = 0.1) +
  labs(x = "House age (years)", y = "Density")
```

Austin house ages: newer construction on the left, the older building boom on the right. Two peaks often mean two subpopulations --- investigate before summarizing with a single number.

## Outliers

:::: {.columns}

::: {.column width="62%"}
::: {.incremental}
- Not everything extreme is an error: the \$800K+ sales in 78752 are real, but a \$13M listing in this dataset was a typo
- The 1.5 × IQR box-plot rule flags potential outliers; it is not a substitute for judgment
:::
:::

::: {.column width="38%"}
```{r}
#| label: viz-boxplot-price
#| fig-width: 3
#| fig-height: 4.5
ggplot(austin_zip, aes(y = latest_price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.3, width = 0.4) +
  scale_y_continuous(labels = dollar_format()) +
  labs(y = "Sale price of homes in 78752") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```
:::

::::
