---
title: "Measuring Location and Spread"
subtitle: "TEMBA Statistics 2026"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")

austin_full <- load_austin_full()

austin_zip <- austin_full |>
  filter(zipcode == 78752)

p49 <- austin_full$latest_price[austin_full$zipcode == 78749]
p41 <- austin_full$latest_price[austin_full$zipcode == 78741]

returns_prices <- read_csv("../data/returns/returns_prices.csv",
                           show_col_types = FALSE) |>
  mutate(date = as.Date(date))

# dollar() inside display math: TeX needs the sign escaped
usd <- function(x) sub("^\\$", "\\\\$", dollar(x))
x4 <- head(austin_zip$latest_price, 4)
```

## We're going to play a game

::: {.incremental}
- You have to guess the sale price of a randomly selected house in one of two ZIP codes.
- You can pick the ZIP code.
:::

## Pick your ZIP code

```{r}
#| label: viz-zip-comparison
#| fig-width: 12
#| fig-height: 4.5
zip_data <- austin_full |>
  filter(zipcode %in% c(78749, 78741)) |>
  mutate(zipcode = factor(zipcode))

p_box <- ggplot(zip_data, aes(x = zipcode, y = latest_price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "darkred") +
  scale_y_continuous(labels = dollar_format()) +
  labs(x = "ZIP code", y = "Sale price")

p_hist <- ggplot(zip_data, aes(x = latest_price, fill = zipcode)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, color = "white", alpha = 0.6,
                 position = "identity") +
  geom_rug(aes(color = zipcode), alpha = 0.3, show.legend = FALSE) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Density", fill = "ZIP code")

p_box + p_hist
```

**Which zip, and what's your guess?**

## The mean

::: {.incremental}
- For the `r nrow(austin_zip)` sales in ZIP 78752, the mean price is `r dollar(round(mean(austin_zip$latest_price)))`: add up all `r nrow(austin_zip)` prices and divide by `r nrow(austin_zip)`.
- In notation, $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$.
:::

## Subscripts and summation

:::: {.columns}

::: {.column width="32%"}
```{r}
#| label: table-subscripts
austin_zip |>
  slice(1:4) |>
  transmute(`$i$` = row_number(),
            `$x_i$ (sale price)` = dollar(latest_price)) |>
  knitr::kable(align = "cl")
```
:::

::: {.column width="68%"}
- A subscript labels *which* observation: $x_1$ is the first sale price, $x_2$ the second, $x_i$ "the $i$-th one." With $n$ observations, $i$ runs from 1 to $n$.
- $\sum$ (capital sigma, for "sum") means "add up":
$$
\sum_{i=1}^{4} x_i = x_1 + x_2 + x_3 + x_4 = `r usd(x4[1])` + `r usd(x4[2])` + `r usd(x4[3])` + `r usd(x4[4])` = `r usd(sum(x4))`.
$$
- Put $n$ in place of 4 to add up the whole column; $\frac{1}{n}\sum_{i=1}^{n} x_i$ is that total divided by $n$ --- the mean.
:::

::::

## The median

::: {.incremental}
- The median price in 78752 is `r dollar(round(median(austin_zip$latest_price)))`: the middle value when the prices are sorted, with half the sales below and half above.
- Unlike the mean, the median doesn't care how extreme the extremes are.
:::

## Mean vs. median

```{r}
#| label: viz-price-mean-median
#| fig-width: 8.5
#| fig-height: 4.2
mean_p <- mean(austin_zip$latest_price)
med_p  <- median(austin_zip$latest_price)
peak_p <- max(density(austin_zip$latest_price)$y)

ggplot(austin_zip, aes(x = latest_price)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, fill = "steelblue", color = "white") +
  geom_density(alpha = 0.3) +
  geom_rug(alpha = 0.4) +
  geom_vline(xintercept = med_p, linetype = "dashed", color = "grey30") +
  geom_vline(xintercept = mean_p, color = "darkorange", linewidth = 0.9) +
  annotate("label", x = med_p, y = peak_p * 0.95, label = "median",
           hjust = 1.1, size = 4.5, fill = "white", alpha = 0.9,
           label.size = 0.25, color = "grey30") +
  annotate("label", x = mean_p, y = peak_p * 0.8, label = "mean",
           hjust = -0.15, size = 4.5, fill = "white", alpha = 0.9,
           label.size = 0.25, color = "darkorange") +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Density")
```

The mean sits above the median because a few high-priced homes drag the average up. Report both, and pay attention to the gap --- a large gap signals skewness.

# Measures of spread

## Back to the game

::: {.incremental}
- Your best guess is the mean or the median.
- Whichever you choose, prices in 78741 vary more around that center, so your prediction error will tend to be larger there.
- Measures of spread quantify how *un*predictable a variable is.
:::

## Standard deviation

The **standard deviation** measures the typical distance of an observation from the mean:

$$
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n} (x_i - \bar{x})^2}
$$

::: {.incremental}
- We square the deviations so positive and negative ones don't cancel, then take the square root to get back to the original units.
- For 78749 the standard deviation is `r dollar(round(sd(p49)))`; for 78741 it is `r dollar(round(sd(p41)))` --- about 1.8 times the standard deviation in 78749.
:::

## Variance

::: {.incremental}
- The **variance** is the standard deviation squared, $s^2$.
- Its units --- dollars-squared here --- are impossible to picture, so we report the standard deviation when we want a readable number.
- Variance earns its keep as a building block inside other calculations.
:::

## Standard deviation measures unpredictability

```{r}
#| label: viz-ford-tesla-sd
#| fig-width: 10
#| fig-height: 5
sd_f <- sd(returns_prices$F); sd_t <- sd(returns_prices$TSLA)
return_series <- function(var, title) {
  ggplot(returns_prices, aes(x = date, y = .data[[var]])) +
    geom_line(color = "steelblue", linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(-0.30, 0.30)) +
    labs(x = NULL, y = "Weekly return", title = title)
}
return_hist <- function(var, s) {
  ggplot(returns_prices, aes(x = .data[[var]])) +
    geom_histogram(binwidth = 0.02, boundary = 0, fill = "steelblue", color = "white") +
    geom_rug(alpha = 0.4) +
    annotate("label", x = 0.29, y = Inf, hjust = 1, vjust = 1.3, size = 4,
             label = paste0("s = ", percent(s, accuracy = 0.1)), label.size = 0) +
    scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(-0.30, 0.30)) +
    labs(x = "Weekly return", y = "Count")
}
(return_series("F", "Ford") | return_series("TSLA", "Tesla")) /
  (return_hist("F", sd_f) | return_hist("TSLA", sd_t))
```

## Same center, different spread

::: {.incremental}
- Both stocks averaged close to zero over these two years: `r percent(mean(returns_prices$F), accuracy = 0.1)` a week for Ford, `r percent(mean(returns_prices$TSLA), accuracy = 0.1)` for Tesla.
- Ford's weekly returns have a standard deviation of `r percent(sd_f, accuracy = 0.1)`; Tesla's is `r percent(sd_t, accuracy = 0.1)` --- about `r number(sd_t / sd_f, accuracy = 0.1)` times larger. A typical Tesla week lands that much farther from zero, in either direction.
- If you had to guess next week's return, zero is about as good a guess as any for either stock. With Tesla you would just be wrong by more.
:::

## The interquartile range

::: {.incremental}
- The **IQR** is the distance between the 75th and 25th percentiles, $\text{IQR} = Q3 - Q1$: the width of the middle half of the data.
- It comes to `r dollar(round(IQR(p49)))` in 78749 and `r dollar(round(IQR(p41)))` in 78741.
- Like the median, the IQR ignores how extreme the extremes are.
:::

# Linear transformations

## Changing units

A **linear transformation** multiplies a variable by a constant, adds a constant, or does both: $y = ax + b$. Every value gets the same rescaling and the same shift.

```{r}
#| label: viz-linear-transform
#| fig-width: 10
#| fig-height: 4
p1 <- ggplot(austin_zip, aes(x = latest_price)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_rug(alpha = 0.4) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Price ($)", y = "Count", title = "Dollars")

p2 <- ggplot(austin_zip, aes(x = latest_price / 1000)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_rug(alpha = 0.4) +
  labs(x = "Price (thousands of dollars)", y = "Count", title = "Thousands of dollars")

p1 + p2
```

The shape is identical; only the axis labels change.

## Shifting: mean-centering

```{r}
#| label: viz-linear-transform-2
#| fig-width: 10
#| fig-height: 4
p3 <- ggplot(austin_zip, aes(x = latest_price - mean(latest_price))) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_rug(alpha = 0.4) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Centered price ($)", y = "Count", title = "Dollars")

p4 <- ggplot(austin_zip, aes(x = (latest_price - mean(latest_price)) / 1000)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_rug(alpha = 0.4) +
  labs(x = "Centered price ($000s)", y = "Count", title = "Thousands")

p3 + p4
```

Subtracting the mean shifts the distribution to center on zero; the shape is unchanged.

## What happens to the summaries?

If $y_i = a x_i + b$ for every observation, where $a$ and $b$ are constants, then

$$
\bar{y} = a\bar{x} + b, \qquad \text{median}(y) = a\,\text{median}(x) + b,
$$

$$
s_y = |a|\,s_x, \qquad s^2_y = a^2\,s^2_x.
$$

Adding $b$ shifts the center and leaves the spread alone; multiplying by $a$ scales both the center and the spread (and the spread doesn't care about the sign of $a$).
