---
title: "Categorical Variables"
subtitle: "TEMBA Statistics 2026"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")

lending <- load_lending()

shed <- read_csv("../data/shed/shed_2025.csv", show_col_types = FALSE)
```

## The data: LendingClub loans

- The `lending` data records `r comma(nrow(lending))` peer-to-peer loans issued between 2008 and 2012, each with a final outcome.
- `purpose` records why the borrower took the loan (14 categories).
- `default` records whether the loan was `Charged Off` --- the borrower stopped paying and the balance was written off --- or `Fully Paid`.

## Kinds of categorical variables

::: {.incremental}
- A **categorical variable** records which of several groups an observation belongs to.
- A **nominal** variable's categories have no natural order: loan `purpose`.
- An **ordinal** variable's categories fall in a natural order, but the spacing between them is not measurable: loan `grade`, from `A` down to `G`.
- A **binary** variable has exactly two categories, so it can be nominal or ordinal: did the loan default?
:::

## Frequency tables {.smaller}

A **frequency table** counts how many observations fall in each category; dividing by the total gives the **proportion**.

::: {style="font-size: 0.62em;"}
```{r}
#| label: table-purpose-counts
lending |>
  count(purpose, name = "Count") |>
  arrange(desc(Count)) |>
  mutate(Proportion = paste0(number(Count / sum(Count), accuracy = 0.001),
                             " (", percent(Count / sum(Count), accuracy = 0.1), ")")) |>
  rename(Purpose = purpose) |>
  knitr::kable(align = c("l", "r", "r"))
```
:::

## Bar charts, dot plots, and pie charts

::: {.incremental}
- A **bar chart** or **dot plot** shows each category's count or proportion as a length or position on a common axis. Differences of a percentage point or two stay visible.
- A **pie chart** shows the same numbers as angles and areas. Two slices a few percentage points apart look the same, and ranking several mid-sized slices is harder still.
- With more than a handful of categories, or when comparing them is the point, use bars or dots.
:::

## The distribution of loan purpose

```{r}
#| label: viz-purpose-dist
#| fig-width: 9
#| fig-height: 4.6
lending |>
  count(purpose) |>
  mutate(prop = n / sum(n),
         purpose = fct_reorder(purpose, prop)) |>
  ggplot(aes(x = purpose, y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Proportion of loans")
```

## The same distribution as a dot plot

```{r}
#| label: viz-purpose-dot
#| fig-width: 9
#| fig-height: 4.6
lending |>
  count(purpose) |>
  mutate(prop = n / sum(n),
         purpose = fct_reorder(purpose, prop)) |>
  ggplot(aes(x = prop, y = purpose)) +
  geom_point(color = "steelblue", size = 3) +
  scale_x_continuous(limits = c(0, NA)) +
  labs(x = "Proportion of loans", y = NULL)
```

## Rating national economic conditions {.smaller}

Each fall the Federal Reserve's [Survey of Household Economics and Decisionmaking](https://www.federalreserve.gov/consumerscommunities/shed.htm) asks a national sample of U.S. adults to rate economic conditions in the country today. The 2025 wave surveyed `r comma(nrow(shed))` people.

```{r}
#| label: viz-shed-outlook
#| fig-width: 8
#| fig-height: 3.8
shed |>
  count(econ_conditions_country) |>
  mutate(prop = n / sum(n),
         econ_conditions_country = factor(econ_conditions_country,
           levels = c("Poor", "Only fair", "Good", "Excellent"))) |>
  ggplot(aes(x = econ_conditions_country, y = prop)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Proportion of respondents")
```

## Ordinal categories keep their order

::: {.incremental}
- We sorted `purpose` by size because its labels have no order. An ordinal variable's bars stay in their natural order: here, worst to best, left to right.
- The tallest bar, "only fair," sits second from the left; the order tells you where that answer falls on the scale.
- Only `r percent(mean(shed$econ_conditions_country == "Excellent"), accuracy = 0.1)` answered "excellent." Kept at the end of the scale, that bar shows the top of the scale is nearly empty; sorted by height, it would just be the rarest answer.
:::

## Dummy (indicator) variables {.smaller}

A **dummy variable** (or **indicator variable**) equals 1 when an observation belongs to a category and 0 when it does not:

$$
\text{default}_i =
\begin{cases}
1 & \text{if loan } i \text{ was charged off} \\
0 & \text{if loan } i \text{ was fully paid.}
\end{cases}
$$

```{r}
#| label: table-dummy-map
lending |>
  filter(loan_status %in% c("Fully Paid", "Charged Off")) |>
  slice(c(1, 3, 4, 2, 5)) |>
  transmute(Purpose = purpose,
            `Loan status` = loan_status,
            default = default) |>
  knitr::kable(align = c("l", "l", "r"))
```

That single column carries everything the two labels told us, in a form we can add up and average.

## A proportion is the mean of a 0/1 variable

::: {.incremental}
- The mean of the `default` dummy is `r number(mean(lending$default), accuracy = 0.001)`, and `r percent(mean(lending$default), accuracy = 0.1)` of these loans were charged off.
- Take the mean of a 0/1 column and you get back the proportion of 1s. Summing the column counts the 1s, so dividing by $n$ gives their share:
:::

::: {.fragment}
$$
\hat{p} = \frac{1}{n}\sum_{i=1}^{n} x_i = \frac{\text{number of 1s}}{n} = \text{proportion of 1s}
$$

We write a sample proportion as $\hat{p}$, read aloud as "p-hat." It is a sample mean in disguise.
:::

## The spread of a 0/1 variable

Predict a random loan's outcome, knowing only the `r percent(mean(lending$default), accuracy = 0.1)` default rate.

::: {.incremental}
- Your best guess is "no default," and you will be right about `r percent(1 - mean(lending$default), accuracy = 0.1)` of the time. That is a fairly predictable outcome.
- A rarer outcome is easier still: if one loan in a hundred defaulted, guessing "no default" every time would be right 99% of the time.
- Where does prediction get hard? At a coin flip: if half of loans defaulted, no guessing rule would beat 50-50.
- So the spread of a 0/1 variable should be smallest near 0 or 1 and largest in the middle.
:::

## Variance of a 0/1 variable

The variance of a 0/1 variable depends on $\hat{p}$ alone:

$$
s^2 \approx \hat{p}\,(1 - \hat{p}),
$$

the chance a randomly selected observation is a 1 times the chance it is a 0.

::: {.incremental}
- Largest at $\hat{p} = 0.5$: a coin flip.
- Shrinks toward zero as $\hat{p}$ approaches 0 or 1: the outcome is nearly certain.
:::

## Variance as a function of $\hat{p}$

```{r}
#| label: viz-variance-parabola
#| fig-width: 8
#| fig-height: 4.2
p_default <- mean(lending$default)

curve_df <- tibble(p = seq(0, 1, length.out = 200),
                   variance = p * (1 - p))

ggplot(curve_df, aes(x = p, y = variance)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey60") +
  annotate("point", x = p_default, y = p_default * (1 - p_default),
           color = "darkred", size = 3) +
  annotate("text", x = p_default, y = p_default * (1 - p_default),
           label = "Loan default", hjust = -0.12, vjust = 1.6, color = "darkred") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = expression(paste("Proportion  ", hat(p))),
       y = expression(paste("Variance  ", hat(p), "(1 - ", hat(p), ")")))
```

At $\hat{p}$ = `r number(p_default, accuracy = 0.001)` the variance is `r number(p_default * (1 - p_default), accuracy = 0.001)`, well below its maximum of 0.25: defaults are fairly predictable. The standard deviation is `r number(sqrt(p_default * (1 - p_default)), accuracy = 0.001)`.
