---
title: "Categorical Variables"
subtitle: "TEMBA Statistics 2026"
---

```{r}
#| label: setup
#| include: false
source("_setup.R")

lending <- load_lending()
```

## The data: LendingClub loans

- The `lending` data records `r comma(nrow(lending))` peer-to-peer loans issued between 2008 and 2012, each with a final outcome.
- `purpose` records why the borrower took the loan (14 categories).
- `default` records whether the loan was `Charged Off` --- the borrower stopped paying and the balance was written off --- or `Fully Paid`.

## Kinds of categorical variables

::: {.incremental}
- A **nominal** variable has categories with no natural order: loan `purpose`.
- An **ordinal** variable has ordered categories, but the spacing between them is not meaningful: loan `grade`, from `A` down to `G`.
- A **binary** variable has exactly two categories: did the loan default?
:::

## Frequency tables {.smaller}

A **frequency table** counts how many observations fall in each category; dividing by the total gives the **proportion**.

```{r}
#| label: table-purpose-counts
purpose_counts <- lending |>
  count(purpose, name = "n") |>
  arrange(desc(n))

bind_rows(
  purpose_counts |> slice_head(n = 7),
  purpose_counts |> slice(-(1:7)) |>
    summarise(purpose = paste0("everything else (", n(), " purposes)"),
              n = sum(n))
) |>
  mutate(Proportion = paste0(number(n / sum(n), accuracy = 0.001),
                             " (", percent(n / sum(n), accuracy = 0.1), ")"),
         Count = comma(n)) |>
  select(Purpose = purpose, Count, Proportion) |>
  knitr::kable(align = c("l", "r", "r"))
```

Two categories cover more than two-thirds of the loans; the rest is a long tail.

## Dummy (indicator) variables {.smaller}

A **dummy variable** (or **indicator variable**) equals 1 when an observation belongs to a category and 0 when it does not:

$$
\text{default}_i =
\begin{cases}
1 & \text{if loan } i \text{ was charged off} \\
0 & \text{if loan } i \text{ was fully paid.}
\end{cases}
$$

```{r}
#| label: table-dummy-map
lending |>
  filter(loan_status %in% c("Fully Paid", "Charged Off")) |>
  slice(c(1, 3, 4, 2, 5)) |>
  transmute(Purpose = purpose,
            `Loan status` = loan_status,
            default = default) |>
  knitr::kable(align = c("l", "l", "r"))
```

That single column is something we can add up and average.

## A proportion is the mean of a 0/1 variable

::: {.incremental}
- The default rate in the lending data is `r percent(mean(lending$default), accuracy = 0.1)`.
- The mean of the `default` dummy is `r number(mean(lending$default), accuracy = 0.001)` --- the same number.
- That's no accident. Summing a column of 0s and 1s counts the 1s, so dividing by $n$ gives the share of 1s:
:::

::: {.fragment}
$$
\hat{p} = \frac{1}{n}\sum_{i=1}^{n} x_i = \frac{\text{number of 1s}}{n} = \text{proportion of 1s}
$$

We write a sample proportion as $\hat{p}$, read aloud as "p-hat."
:::

## The spread of a 0/1 variable

Predict a random loan's outcome, knowing only the `r percent(mean(lending$default), accuracy = 0.1)` default rate.

::: {.incremental}
- Your best guess is "no default," and you will be right about `r percent(1 - mean(lending$default), accuracy = 0.1)` of the time. That is a fairly predictable outcome.
- Where does prediction get hard? At a coin flip: if half of loans defaulted, no guess would beat 50-50.
- So the spread of a 0/1 variable should be smallest near 0 or 1 and largest in the middle.
:::

## Variance of a 0/1 variable

A 0/1 variable can only deviate from its mean two ways: a 1 sits $(1 - \hat{p})$ above it, and a 0 sits $\hat{p}$ below it. The variance formula collapses to

$$
s^2 = \frac{n}{n-1}\,\hat{p}\,(1 - \hat{p}) \approx \hat{p}\,(1 - \hat{p}).
$$

In words: the variance of a yes/no variable is the chance of a yes times the chance of a no.

## Variance as a function of $\hat{p}$

```{r}
#| label: viz-variance-parabola
#| fig-width: 8
#| fig-height: 4.2
p_default <- mean(lending$default)

curve_df <- tibble(p = seq(0, 1, length.out = 200),
                   variance = p * (1 - p))

ggplot(curve_df, aes(x = p, y = variance)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey60") +
  annotate("point", x = p_default, y = p_default * (1 - p_default),
           color = "darkred", size = 3) +
  annotate("text", x = p_default, y = p_default * (1 - p_default),
           label = "Loan default", hjust = -0.12, vjust = 1.6, color = "darkred") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = expression(paste("Proportion  ", hat(p))),
       y = expression(paste("Variance  ", hat(p), "(1 - ", hat(p), ")")))
```

Spread is zero at the extremes, where the outcome is certain, and largest at the coin flip. The default proportion lands well down the left slope: a loan is far more likely to be repaid than not, so its outcome is comparatively predictable.
