---
title: "Introduction to Probability and Sampling"
subtitle: "TEMBA Statistics 2026"
fig-align: center
---

```{r}
#| label: setup
#| include: false
source("_setup.R")

lending_subs <- load_lending_subs()
shed <- load_shed()

sb_rate <- mean(lending_subs$default[lending_subs$purpose == "small_business"])
dc_rate <- mean(lending_subs$default[lending_subs$purpose == "debt_consolidation"])
shed_good_rate <- mean(shed$econ_conditions_country == "Good")
```

## You're funding one of two loans

The loans have the same amount, terms, and borrower credit history; only their purposes differ. The historical data are:

```{r}
#| label: table-two-loan-defaults
lending_subs |>
  filter(purpose %in% c("small_business", "debt_consolidation")) |>
  group_by(purpose) |>
  summarise(n = n(), default_rate = mean(default)) |>
  arrange(desc(default_rate)) |>
  mutate(n = comma(n),
         default_rate = percent(default_rate, accuracy = 0.1)) |>
  rename(Purpose = purpose, Loans = n, `Default rate` = default_rate) |>
  knitr::kable(align = c("l", "r", "r"))
```

**Which loan would you fund, and what's the chance it defaults?**

## You did a number of things at once

::: {.incremental}
- You probably picked the debt-consolidation loan and said there's a `r percent(dc_rate, accuracy = 0.1)` chance of default.
- You decided the historical data are *representative* of future outcomes.
- You constructed a rule for assigning probabilities to uncertain outcomes --- a **probability model** --- estimated its **parameters** (the two default probabilities), and used it to make a decision.
:::

## Two interpretations of probability

:::: {.columns}

::: {.column width="48%"}
**Long-run frequency**

If we could fund many comparable debt-consolidation loans, about `r percent(dc_rate, accuracy = 0.1)` would default.
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
**Subjective probability**

For this one new loan, `r percent(dc_rate, accuracy = 0.1)` expresses our current assessment given what we know.
:::

::::

Both interpretations follow the same probability rules.

## FedWatch probabilities are market-implied {.smaller}

The [CME FedWatch](https://www.cmegroup.com/markets/interest-rates/cme-fedwatch-tool.html) converts futures prices into probabilities for the federal funds target range. These probabilities were recorded on July 9, 2026.

:::: {.columns}

::: {.column width="48%"}
**July 29 meeting**

```{r}
#| label: table-cme-intro-july
cme_july |>
  complete(
    bucket = c("3.50--3.75%", "3.75--4.00%", "4.00--4.25%"),
    fill = list(prob = 0)
  ) |>
  arrange(bucket) |>
  transmute(`Target range` = bucket,
            Probability = number(prob, accuracy = 0.001)) |>
  knitr::kable(align = c("l", "r"))
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
**September 16 meeting**

```{r}
#| label: table-cme-intro-sept
cme_sept |>
  complete(
    bucket = c("3.50--3.75%", "3.75--4.00%", "4.00--4.25%"),
    fill = list(prob = 0)
  ) |>
  arrange(bucket) |>
  transmute(`Target range` = bucket,
            Probability = number(prob, accuracy = 0.001)) |>
  knitr::kable(align = c("l", "r"))
```
:::

::::

These are market-implied model probabilities for two one-off meetings, written as probability distributions.

## Probability describes data we don't have

::: {.incremental}
- Histograms, densities, ECDFs, and frequency tables describe distributions of **observed** data.
- We usually want to know about data we *don't* have. Examples include the sale price of a new listing, the Fed funds rate after the next meeting, and sentiment among all U.S. adults.
- A **probability distribution** assigns probabilities to every possible outcome of a random or uncertain process. It describes future or unobserved data.
:::

## Historical rates and model probabilities

```{r}
#| label: table-loan-default-model
tibble(
  Purpose = c("Small business", "Debt consolidation"),
  `Historical default rate` = percent(c(sb_rate, dc_rate), accuracy = 0.1),
  `Pr(default)` = number(c(sb_rate, dc_rate), accuracy = 0.001),
  `Pr(repaid)` = number(c(1 - sb_rate, 1 - dc_rate), accuracy = 0.001)
) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

A **probability model** approximates a real process. Using historical rates as future probabilities assumes that these loans represent future loans; the model is useful only if it captures the features relevant to the decision.

## Historical churn as a probability model {.smaller}

A software company observed 90 cancellations among 1,200 annual subscriptions last year.

1. What was the historical churn rate?
2. If 200 otherwise similar subscriptions renew this year, how many cancellations does that model predict?

::: {.fragment}
$$
\frac{90}{1{,}200}=0.075, \qquad 200(0.075)=15.
$$

::: {.callout-warning}
The company raised its annual subscription price by 15% this year. Does last year's churn rate still provide a credible probability model?
:::
:::

## Sampling, two ways

::: {.incremental}
- A **sample** is a collection of observations.
- **Finite-population sampling** selects observations from a population. The SHED survey sampled `r comma(nrow(shed))` adults, and `r percent(shed_good_rate, accuracy = 0.1)` rated national economic conditions **Good**.
- **Probability-model sampling** records repeated outcomes from a random process, such as rolls of a four-sided die.
- In either setting, the sample must represent what we are trying to learn about.
:::

## Survey responses and die rolls {.smaller}

Suppose exactly 25% of U.S. adults would rate national economic conditions **Good**. Sampling adults then has the same probability structure as rolling a fair four-sided die.

| | Survey sample | Four-sided die rolls |
|:--|:--|:--|
| One trial | Randomly sample one adult | Roll the die once |
| Outcome of interest | Rates conditions "Good" | Roll is 4 |
| Chance under the model | 0.25 | 0.25 |
| Sample size | `r comma(nrow(shed))` adults | About 13,000 rolls |

In SHED, `r percent(shed_good_rate, accuracy = 0.1)` of respondents rated conditions **Good**.

## Representativeness depends on the target {.smaller}

:::: {.columns}

::: {.column width="48%"}
**Team A**

The team randomly samples 400 loyalty-program members. In the sample, 36% use curbside pickup.
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
**Team B**

The team randomly samples 400 customers who clicked a curbside-pickup email. In the sample, 36% use curbside pickup.
:::

::::

Which population can each result credibly describe? Does either result represent all customers?

::: {.fragment}
Each team observed 144 curbside-pickup users. Team A can represent loyalty-program members; Team B can represent email clickers. Neither sampling frame covers all customers.
:::
