---
title: "Joint, Marginal, and Conditional Distributions"
subtitle: "TEMBA Statistics 2026"
format:
  revealjs:
    chalkboard: true
    embed-resources: false
execute:
  freeze: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
library(tidyverse)
library(scales)

# Figures render at their intrinsic size (fig-width x 96px), capped by the
# ~702px content column -- they are never upscaled to fill it. So axis text
# lands at its true point size, and base_size, not fig-width, is what makes it
# legible next to 17px body text. At 13, tick labels are ~14px.


lending_subs <- read_csv("../data/lending_club/data/df_subs.csv",
                         show_col_types = FALSE) |>
  mutate(default = as.integer(loan_status == "Charged Off"))

# Historical LendingClub counts and the resulting probability model for a
# future loan. All loan calculations below derive from this single crosstab.
loan_counts <- lending_subs |>
  count(purpose, loan_status) |>
  pivot_wider(names_from = loan_status, values_from = n, values_fill = 0) |>
  transmute(purpose,
            repaid = `Fully Paid`,
            default = `Charged Off`,
            Total = repaid + default) |>
  arrange(desc(Total))

loan_n <- sum(loan_counts$Total)

loan_joint <- loan_counts |>
  transmute(purpose,
            repaid = repaid / loan_n,
            default = default / loan_n)

loan_small <- loan_counts |> filter(purpose == "small_business")
loan_debt <- loan_counts |> filter(purpose == "debt_consolidation")
loan_home <- loan_counts |> filter(purpose == "home_improvement")
loan_default_n <- sum(loan_counts$default)

stopifnot(
  isTRUE(all.equal(sum(loan_joint$repaid) + sum(loan_joint$default), 1)),
  isTRUE(all.equal(rowSums(loan_joint[-1]),
                   loan_counts$Total / loan_n)),
  isTRUE(all.equal(sum(loan_joint$default),
                   loan_default_n / loan_n)),
  isTRUE(all.equal(
    loan_small$default / loan_small$Total,
    (loan_small$default / loan_n) / (loan_small$Total / loan_n)
  ))
)


p3 <- function(x) number(x, accuracy = 0.001)
p4 <- function(x) number(x, accuracy = 0.0001)
```

## Loan purpose and default

The [LendingClub extract](../notes/appendix_data.html#sec-data-lendingclub-kaggle) contains `r comma(loan_n)` completed loans with these three purposes.

```{r}
#| label: table-purpose-default-counts
#| fig-width: 9
#| fig-height: 3.9
purpose_counts <- lending_subs |>
  count(purpose, loan_status) |>
  pivot_wider(names_from = loan_status, values_from = n, values_fill = 0) |>
  mutate(Total = `Fully Paid` + `Charged Off`) |>
  arrange(desc(Total))

purpose_counts |>
  bind_rows(summarise(purpose_counts, purpose = "**Total**",
                      across(where(is.numeric), sum))) |>
  mutate(across(where(is.numeric), comma)) |>
  mutate(purpose = str_to_sentence(str_replace_all(purpose, "_", " "))) |>
  rename(Purpose = purpose) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

Which purpose has the greatest default rate? Which denominator would you use?

## A cross-tabulation records combinations

A **cross-tabulation** counts observations in every combination of two categories.

```{r}
#| label: table-joint-margins
#| fig-width: 9
#| fig-height: 3.9
loan_counts |>
  transmute(Purpose = str_to_sentence(str_replace_all(purpose, "_", " ")), Repaid = repaid, Default = default, Total) |>
  bind_rows(summarise(loan_counts, Purpose = "Total", Repaid = sum(repaid), Default = sum(default), Total = sum(Total))) |>
  knitr::kable(format.args = list(big.mark = ","))
```

The interior cells describe pairs of outcomes. The margins summarize each variable on its own.

## Joint and marginal proportions

Both use the full sample as their denominator.

```{r}
#| label: table-joint-divisions
#| fig-width: 9
#| fig-height: 3.9
tibble(Question = c("Small business and default", "Small business, either outcome"),
       Calculation = c(paste(comma(loan_small$default), "/", comma(loan_n)), paste(comma(loan_small$Total), "/", comma(loan_n))),
       Proportion = number(c(loan_small$default, loan_small$Total)/loan_n, accuracy = 0.0001)) |>
  knitr::kable()
```

A **joint proportion** describes a combination. A **marginal proportion** describes one category, summing over the other variable.

## Conditional proportions select a group

Among small-business loans, what proportion defaulted?

```{r}
#| label: table-conditional-small
#| fig-width: 9
#| fig-height: 3.9
tibble(Outcome = c("Repaid", "Default", "Total"),
       Count = comma(c(loan_small$repaid, loan_small$default, loan_small$Total)),
       `Conditional proportion` = number(c(loan_small$repaid, loan_small$default, loan_small$Total)/loan_small$Total, accuracy = 0.001)) |>
  knitr::kable()
```

The **conditional distribution** divides each count by the selected group's total. Its proportions sum to one.

## Reversing the condition changes the denominator

```{r}
#| label: table-reversed-condition
#| fig-width: 9
#| fig-height: 3.9
tibble(Question = c("Default, given small business", "Small business, given default"),
       Calculation = c(paste(comma(loan_small$default), "/", comma(loan_small$Total)), paste(comma(loan_small$default), "/", comma(loan_default_n))),
       Proportion = number(c(loan_small$default/loan_small$Total, loan_small$default/loan_default_n), accuracy = 0.001)) |>
  knitr::kable()
```

Name the population described by each row. Why can the same numerator answer two different questions?

## Hybrid work and attrition

[Bloom, Han, and Liang](../notes/appendix_data.html#sec-data-wfh-bloom) randomly assigned 1,612 employees to two work schedules.

| Assigned schedule | Stayed | Left | Total |
|---|---:|---:|---:|
| Control | 761 | 59 | 820 |
| Hybrid | 754 | 38 | 792 |
| Total | 1,515 | 97 | 1,612 |

How does attrition compare between the two schedules?

## Which number belongs in the briefing?

Three statements use the same 38 employees:

1. About 2.4% of all employees were assigned hybrid work and left.
2. About 4.8% of hybrid employees left.
3. About 39.2% of employees who left were assigned hybrid work.

Are any numerically wrong? Which answers the comparison question?

**In pairs:** prepare a two-sentence briefing. Include the other schedule and identify both denominators.

## Attrition within each schedule

| Assigned schedule | Stayed | Left | Denominator |
|---|---:|---:|---:|
| Control | 0.9280 | 0.0720 | 820 |
| Hybrid | 0.9520 | 0.0480 | 792 |

All three earlier statements were numerically correct. Only the second described attrition within a schedule.

About 4.8% of hybrid employees left, compared with 7.2% of control employees. Each row is a complete conditional distribution.

## From observed proportions to a probability model

We can use historical proportions as estimated probabilities for a future loan if the historical loans represent that future population.

```{r}
#| label: table-loan-joint
#| fig-width: 9
#| fig-height: 3.9
loan_joint |>
  mutate(purpose = case_when(
           purpose == "debt_consolidation" ~ "Debt consolidation (P = 1)",
           purpose == "home_improvement" ~ "Home improvement (P = 2)",
           purpose == "small_business" ~ "Small business (P = 3)"
         ),
         across(where(is.numeric), \(x) number(x, accuracy = 0.001))) |>
  rename(Purpose = purpose, `Repaid (D = 0)` = repaid,
         `Default (D = 1)` = default) |>
  knitr::kable(align = c("l", "r", "r"))
```

The six interior probabilities are nonnegative and sum to one. The numbers now describe possible outcomes for a future loan.

## Conditioning a probability table

$$
\Pr(A\mid B)=\frac{\Pr(A\text{ and }B)}{\Pr(B)},\qquad \Pr(B)>0.
$$

Keep the outcomes consistent with the condition, then divide by their total probability.

For the loan model, dividing the small-business/default joint probability by the small-business margin gives the same rate as dividing their counts.

## Complete a renewal model

| Current tier | Renews | Does not renew | Total |
|---|---:|---:|---:|
| Basic | 0.42 | 0.28 | 0.70 |
| Premium | 0.24 | ? | 0.30 |
| Total | ? | ? | 1.00 |

Find the missing cell and column margins. Then find the renewal probability given Premium.

An analyst calls 0.24 the Premium renewal rate. Is that the right interpretation?

## Renewal probabilities

| Current tier | Renews | Does not renew | Total |
|---|---:|---:|---:|
| Basic | 0.42 | 0.28 | 0.70 |
| Premium | 0.24 | 0.06 | 0.30 |
| Total | 0.66 | 0.34 | 1.00 |

The Premium renewal probability is $0.24/0.30=0.80$.

The joint probability 0.24 describes being Premium **and** renewing. Its denominator is the entire customer population.
