---
title: "Independence"
subtitle: "TEMBA Statistics 2026"
format:
  revealjs:
    chalkboard: true
    embed-resources: false
execute:
  freeze: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
library(scales)

loans_ind <- read.csv("../data/lending_club/data/df_subs.csv")
purpose_ind <- c("small_business", "debt_consolidation", "home_improvement")
loan_counts_ind <- table(factor(loans_ind$purpose, levels = purpose_ind),
                         factor(loans_ind$loan_status,
                                levels = c("Charged Off", "Fully Paid")))
loan_n_ind <- sum(loan_counts_ind)
loan_joint_ind <- loan_counts_ind / loan_n_ind
loan_default_ind <- sum(loan_counts_ind[, 1]) / loan_n_ind
loan_rates_ind <- loan_counts_ind[, 1] / rowSums(loan_counts_ind)
stopifnot(loan_n_ind == nrow(loans_ind), loan_n_ind == 58456,
          all(rowSums(loan_counts_ind) == c(3191, 49409, 5856)),
          abs(sum(loan_joint_ind) - 1) < 1e-12,
          abs(sum(loan_joint_ind[, 1]) - loan_default_ind) < 1e-12)

# July 9, 2026 published CME histories; the supported moves determine zeros.
july_ind <- c(0.754138, 0.245862)
sept_ind <- c(0.376530, 0.500363, 0.123107)
cme_ind <- matrix(c(sept_ind[1], july_ind[1] - sept_ind[1], 0,
                    0, july_ind[2] - sept_ind[3], sept_ind[3]),
                  nrow = 2, byrow = TRUE,
                  dimnames = list(c("3.50–3.75%", "3.75–4.00%"),
                                  c("3.50–3.75%", "3.75–4.00%", "4.00–4.25%")))
stopifnot(all(cme_ind >= 0), abs(sum(cme_ind) - 1) < 1e-12,
          max(abs(rowSums(cme_ind) - july_ind)) < 1e-12,
          max(abs(colSums(cme_ind) - sept_ind)) < 1e-12)

ad_ind <- c(A = 0.6, B = 0.4)
device_ind <- c(Phone = 0.5, Tablet = 0.2, Computer = 0.3)
model_ind <- outer(ad_ind, device_ind)
stopifnot(abs(sum(model_ind) - 1) < 1e-12,
          max(abs(rowSums(model_ind) - ad_ind)) < 1e-12,
          max(abs(colSums(model_ind) - device_ind)) < 1e-12,
          max(abs(sweep(model_ind, 1, ad_ind, "/") -
                    matrix(device_ind, 2, 3, byrow = TRUE))) < 1e-12,
          max(abs(sweep(model_ind, 2, device_ind, "/") -
                    matrix(ad_ind, 2, 3))) < 1e-12)

shed_ind <- read.csv("../data/shed/shed_2025.csv")
employment_ind <- c("Not working", "Working full-time", "Working part-time")
ratings_ind <- c("Excellent", "Good", "Only fair", "Poor")
shed_counts_ind <- table(factor(shed_ind$employment_status, levels = employment_ind),
                         factor(shed_ind$econ_conditions_country, levels = ratings_ind))
shed_cond_ind <- prop.table(shed_counts_ind, 1)
shed_marg_ind <- colSums(shed_counts_ind) / sum(shed_counts_ind)
shed_gap_ind <- max(abs(sweep(shed_cond_ind, 2, shed_marg_ind, "-")))
stopifnot(sum(shed_counts_ind) == nrow(shed_ind), nrow(shed_ind) == 12934,
          max(abs(rowSums(shed_cond_ind) - 1)) < 1e-12,
          abs(sum(shed_marg_ind) - 1) < 1e-12,
          round(100 * shed_gap_ind, 1) == 1.4)
```

## Does loan purpose change the distribution?

The [LendingClub subset](../notes/appendix_data.html#sec-data-lendingclub-kaggle) gives these historical default rates.

```{r}
#| label: table-ind-loan-rates
#| fig-width: 9
#| fig-height: 3.9
data.frame(Purpose = c("Small business", "Debt consolidation", "Home improvement"),
           Loans = comma(rowSums(loan_counts_ind)),
           `Default rate` = percent(loan_rates_ind, accuracy = 0.1),
           check.names = FALSE) |>
  knitr::kable(align = c("l", "r", "r"), row.names = FALSE)
```

The overall rate is `r percent(loan_default_ind, accuracy = 0.1)`. Which purposes change our assessment of default?

## Independence means unchanged distributions

Two variables are **independent** when learning one does not change the distribution of the other.

$$
\Pr(X=x\mid Y=y)=\Pr(X=x).
$$

For discrete variables, this must hold for every $x$ and every conditioning value $y$ with positive probability. The same property holds in the reverse direction.

## Independent joint probabilities factor

For discrete variables, independence is equivalent to

$$
\Pr(X=x,Y=y)=\Pr(X=x)\Pr(Y=y)
$$

for every pair of outcomes.

Each interior cell must equal its row margin times its column margin. One failed equality establishes dependence.

## Two promotional offers

A system assigns free shipping independently of a discount.

| Shipping offer | Discount | No discount | Total |
|---|---:|---:|---:|
| Free shipping | ? | ? | 0.25 |
| No free shipping | ? | ? | 0.75 |
| Total | 0.40 | 0.60 | 1.00 |

Complete the model. What should the probability of receiving both offers be?

## Can this model meet the specification?

An engineer proposes probability **0.20** for receiving both offers, keeping the same margins.

| Shipping offer | Discount | No discount | Total |
|---|---:|---:|---:|
| Free shipping | 0.20 | ? | 0.25 |
| No free shipping | ? | ? | 0.75 |
| Total | 0.40 | 0.60 | 1.00 |

Give two verdicts: can this be a valid model, and can it satisfy independence? Complete the remaining cells to justify your answer.

## The independent model

| Shipping offer | Discount | No discount | Total |
|---|---:|---:|---:|
| Free shipping | 0.10 | 0.15 | 0.25 |
| No free shipping | 0.30 | 0.45 | 0.75 |
| Total | 0.40 | 0.60 | 1.00 |

The conditional discount probabilities are $0.10/0.25=0.40$ and $0.30/0.75=0.40$.

Learning the shipping offer leaves the discount distribution unchanged.

## A valid model can be dependent

| Shipping offer | Discount | No discount | Total |
|---|---:|---:|---:|
| Free shipping | 0.20 | 0.05 | 0.25 |
| No free shipping | 0.20 | 0.55 | 0.75 |
| Total | 0.40 | 0.60 | 1.00 |

All cells are nonnegative and sum to one, with the required margins.

But $0.20\ne0.25(0.40)$. The conditional discount probabilities are 0.80 with free shipping and about 0.267 without it.

## One cell can disprove independence

| | Purchase | No purchase | Total |
|---|---:|---:|---:|
| Store visit | 0.18 | 0.22 | 0.40 |
| No store visit | 0.12 | 0.48 | 0.60 |
| Total | 0.30 | 0.70 | 1.00 |

Purchases can also occur online. Find one numerical comparison that establishes dependence.

## The product check fails

Independence would require

$$
\Pr(\text{visit and purchase})=0.40(0.30)=0.12.
$$

The model assigns 0.18 to that cell, so store visits and purchases are dependent.

We do not need to check the remaining cells once one equality fails.

## Similar observed distributions

The [2025 SHED survey](../notes/appendix_data.html#sec-data-shed) records ratings of national economic conditions. These percentages are unweighted.

```{r}
#| label: table-ind-shed
#| fig-width: 9
#| fig-height: 3.9
shed_display_ind <- rbind(shed_cond_ind, Overall = shed_marg_ind)
shed_display_ind[] <- percent(shed_display_ind, accuracy = 0.1)
knitr::kable(shed_display_ind, align = rep("r", 4))
```

The largest difference from an overall category share is `r number(100*shed_gap_ind, accuracy = 0.1)` percentage points. Does similarity establish exact population independence?

## Sample similarity and model independence

Sample proportions generally differ even when a population model has independence. Rounding can also conceal differences.

The SHED table suggests that employment status tells us relatively little about these ratings in this extract. It does not establish exact independence in the population.

For quantitative variables, independence likewise requires the entire conditional distribution to stay unchanged: its location, spread, and shape.
