---
title: "Visualizing and Summarizing Two Quantitative Variables"
subtitle: "TEMBA Statistics 2026"
format:
  revealjs:
    chalkboard: true
    embed-resources: false
execute:
  freeze: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
# Companion: ../notes/prob_08_quantitative-quantitative.qmd (Chapter 13).

austin <- read_csv("../data/austin_houses/austin_houses_subset.csv",
                   show_col_types = FALSE)

returns <- read_csv("../data/returns/returns.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date))

load1 <- read_csv("../data/ercot/ercotproject/Data/load_1.csv",
                  show_col_types = FALSE)
load2 <- read_csv("../data/ercot/ercotproject/Data/load_2.csv",
                  show_col_types = FALSE)
load3 <- read_csv("../data/ercot/ercotproject/Data/load_3.csv",
                  show_col_types = FALSE)
ercot <- bind_rows(load1, load2, load3) |>
  filter(LOCATION == "North Central (DFW)") |>
  mutate(datetime = parse_date_time(DATEHOUR, orders = c("ymd HM", "ymd HMS")),
         year = year(datetime),
         date = as.Date(datetime)) |>
  filter(year >= 2016, year <= 2020) |>
  summarize(LOAD = mean(LOAD, na.rm = TRUE),
            date = first(date),
            .by = c(datetime, LOCATION))
rm(load1, load2, load3)

weather <- read_csv("../data/ercot/ercotproject/Data/weather_ncent.csv",
                    show_col_types = FALSE) |>
  mutate(datetime = parse_date_time(DATEHOUR, orders = c("ymd HM", "ymd HMS")),
         year = year(datetime)) |>
  filter(year >= 2016, year <= 2020) |>
  summarize(HourlyDryBulbTemperature = mean(HourlyDryBulbTemperature, na.rm = TRUE),
            .by = c(datetime, LOCATION))

ercot_temp <- ercot |>
  inner_join(weather, by = c("datetime", "LOCATION")) |>
  filter(!is.na(HourlyDryBulbTemperature), !is.na(LOAD))

ercot_daily <- ercot_temp |>
  summarize(temp = mean(HourlyDryBulbTemperature, na.rm = TRUE),
            load = mean(LOAD, na.rm = TRUE),
            .by = date)
rm(weather)
```

## Price and living area

Each point is one [Austin home sale](../notes/appendix_data.html#sec-data-housing-kaggle).

```{r}
#| label: viz-austin-price-sqft
#| fig-width: 9
#| fig-height: 3.7
ggplot(austin, aes(x = living_area_sq_ft, y = latest_price)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  scale_x_continuous(labels = comma_format()) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Living area (sq ft)", y = "Sale price")
```

Describe the trend and the spread of prices among similarly sized homes.

## A smooth curve summarizes nearby outcomes

[DFW daily electricity demand and temperature](../notes/appendix_data.html#sec-data-ercot-shaffer), 2016–2020.

```{r}
#| label: viz-ercot-load-temp
#| fig-width: 9
#| fig-height: 3.7
# Windowed running-line smoother. Keep h and the grid IDENTICAL to
# working/loess_animation.R, which animates this exact construction.
h <- 8
x_grid <- seq(min(ercot_daily$temp) + h, max(ercot_daily$temp) - h, length.out = 80)

local_fit <- function(x0) {
  window <- filter(ercot_daily, abs(temp - x0) <= h)
  fit <- lm(load ~ temp, data = window)
  predict(fit, newdata = tibble(temp = x0))
}
smooth_df <- tibble(temp = x_grid, load = map_dbl(x_grid, local_fit))
load_at_50 <- unname(local_fit(50))

ggplot(ercot_daily, aes(x = temp, y = load)) +
  geom_point(alpha = 0.3, size = 0.8, color = "grey50") +
  geom_line(data = smooth_df, color = "steelblue", linewidth = 1.1) +
  scale_y_continuous(labels = comma_format()) +
  labs(x = expression("Temperature ("*degree*"F)"), y = "Load (MW)")
```

At 50°F, the curve is about `r comma(round(load_at_50, -3))` MW. What does the surrounding scatter tell us?

## Conditional location and spread

The smoother describes average demand among days with nearby temperatures.

The scatter around it describes variation among individual days at those temperatures.

Knowing the temperature narrows our assessment of demand, but does not determine the load on a particular day.

## A smoother also helps with housing prices

```{r}
#| label: viz-austin-price-sqft-smooth
#| fig-width: 9
#| fig-height: 4.1
ggplot(austin, aes(x = living_area_sq_ft, y = latest_price)) +
  geom_point(alpha = 0.25, color = "grey50") +
  geom_smooth(method = "loess", span = 0.75, se = FALSE,
              color = "steelblue", linewidth = 1) +
  scale_x_continuous(labels = comma_format()) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Living area (sq ft)", y = "Sale price")
```

Fewer large homes determine the curve at the right. Small bends there need not describe a stable population pattern.

## A roughly linear relationship

Each point pairs one week's [Tesla and S&P 500 returns](../notes/appendix_data.html#sec-data-returns-2025).

```{r}
#| label: viz-tsla-market
#| fig-width: 9
#| fig-height: 3.7
ggplot(returns, aes(x = market, y = TSLA)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "S&P 500 weekly return", y = "Tesla weekly return")
```

The broad trend is upward and roughly straight, with considerable variation around it.

## Correlation summarizes linear association

The sample **correlation**, $r$, ranges from $-1$ to $1$ and describes the direction and strength of linear association.

- Positive values describe upward association; negative values describe downward association.
- Values nearer $1$ or $-1$ describe stronger linear association.
- At $1$ or $-1$, all points lie on a straight line.

Correlation has no units. Both variables must vary for it to be defined.

## Different population correlations

```{r}
#| label: viz-corr-panels
#| fig-width: 9.5
#| fig-height: 5.5
set.seed(42)
rs <- c(1, 0.75, 0.5, 0.25, 0, -0.25, -0.5, -0.75, -1)
n <- 500

corr_panels <- map_dfr(rs, function(r) {
  x <- rnorm(n)
  y <- r * x + sqrt(1 - r^2) * rnorm(n)
  tibble(x = x, y = y, panel = match(r, rs),
         label = number(r, accuracy = 0.01))
}) |>
  mutate(label = factor(label, levels = unique(label)))

ggplot(corr_panels, aes(x = x, y = y)) +
  geom_point(size = 0.4, alpha = 0.4, color = "steelblue") +
  facet_wrap(~label, ncol = 3) +
  coord_equal() +
  theme_minimal(base_size = 20) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(x = NULL, y = NULL)
```

Each cloud contains 500 observations. The labels describe the generating models; sample correlations fluctuate around them.

## What would you ask to see?

A report gives essentially the same correlation, $r=0.82$, for four datasets.

The author proposes using the same description of the relationship for all four.

What else would you ask to see before approving the summary?

## Anscombe's quartet

```{r}
#| label: viz-anscombe
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
data(anscombe)

anscombe_long <- tibble(
  set = rep(1:4, each = nrow(anscombe)),
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
) |>
  mutate(
    set = factor(set, levels = 1:4, labels = paste("Set", 1:4)),
    r = cor(x, y),
    .by = set
  )

label_data <- anscombe_long |>
  distinct(set, r) |>
  mutate(
    x = -Inf,
    y = Inf,
    label = sprintf("r = %.2f", r)
  )

ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(color = "steelblue", alpha = 0.8) +
  geom_label(
    data = label_data,
    aes(x = x, y = y, label = label),
    hjust = -0.1, vjust = 1.2, size = 3.85, fill = "white", alpha = 0.9, label.size = 0.25, inherit.aes = FALSE
  ) +
  facet_wrap(~set, ncol = 2) +
  theme(aspect.ratio = 3 / 4) +
  labs(x = "x", y = "y")
```


## What the four plots add

- The upper-left panel has a roughly linear pattern.
- The upper-right panel follows a curve.
- In the lower-left panel, one point departs from the others' straight pattern.
- In the lower-right panel, one observation accounts for the positive correlation.

A correlation alone cannot distinguish these cases.

## Similar correlations can describe different patterns

```{r}
#| label: viz-corr-linear-only
#| fig-width: 10
#| fig-height: 4.1
p_tsla <- ggplot(returns, aes(x = market, y = TSLA)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  annotate("label", x = -Inf, y = Inf,
           label = sprintf("r = %.2f", cor(returns$market, returns$TSLA)),
           hjust = -0.1, vjust = 1.2, size = 3.5) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "S&P 500 weekly return", y = "Tesla weekly return")

p_ercot <- ggplot(ercot_daily, aes(x = temp, y = load)) +
  geom_point(alpha = 0.3, size = 0.8, color = "steelblue") +
  scale_y_continuous(labels = comma_format()) +
  annotate("label", x = -Inf, y = Inf,
           label = sprintf("r = %.2f", cor(ercot_daily$temp, ercot_daily$load)),
           hjust = -0.1, vjust = 1.2, size = 3.5) +
  labs(x = expression("Temperature ("*degree*"F)"), y = "Load (MW)")

p_tsla | p_ercot
```

Compare the scatter around each trend with the overall variation in its outcome. Which relationship does the correlation describe less well?

## Zero correlation does not establish independence

Correlation measures linear association. A strong curved relationship can have a correlation near zero.

Independence requires the **entire conditional distribution** to stay unchanged. Even exactly zero correlation is insufficient.

A sample correlation near zero also does not establish independence in a population. Inspect the scatterplot.

## Covariance retains the measurement scales

The sample **covariance** combines correlation and standard deviations:

$$
s_{xy}=r\,s_xs_y.
$$

It has the same sign as correlation, but its magnitude depends on the units. For size and price, its units are square feet times dollars.

If prices change from dollars to thousands of dollars, what happens to covariance? What happens to correlation?

## Changing units does not weaken the relationship

| Quantity | Dollars → thousands of dollars |
|---|---|
| Price standard deviation | Divides by 1,000. |
| Covariance of size and price | Divides by 1,000. |
| Correlation of size and price | Stays the same. |

Covariance is not restricted to the interval from $-1$ to $1$. Its numerical scale changes with the variables' scales.

## Covariance in a probability model

For random variables with finite variances,

$$
\operatorname{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)].
$$

Here $\mu_X$ and $\mu_Y$ are the model means. Covariance is the probability-weighted average of the products of deviations.

For positive standard deviations,

$$
\operatorname{Cov}(X,Y)=\operatorname{Corr}(X,Y)\operatorname{SD}(X)\operatorname{SD}(Y).
$$

## A joint model for two meeting outcomes

[CME FedWatch](https://www.cmegroup.com/markets/interest-rates/cme-fedwatch-tool.html), July 9, 2026: a model for July and September target ranges. A midpoint of 3.625 represents 3.50–3.75%.

```{r}
#| label: table-cme-cov-model
#| fig-width: 9
#| fig-height: 3.9
cme_model <- tibble(
  x = c(3.625, 3.625, 3.875, 3.875),
  y = c(3.625, 3.875, 3.875, 4.125),
  p = c(0.376530, 0.377608, 0.122755, 0.123107)
)
mx <- sum(cme_model$x*cme_model$p)
my <- sum(cme_model$y*cme_model$p)
sx <- sqrt(sum(cme_model$p*(cme_model$x-mx)^2))
sy <- sqrt(sum(cme_model$p*(cme_model$y-my)^2))
cov_model <- sum(cme_model$p*(cme_model$x-mx)*(cme_model$y-my))
cor_model <- cov_model/(sx*sy)
stopifnot(abs(sum(cme_model$p)-1)<1e-12,
          abs(sum(cme_model$p[cme_model$x==3.625])-0.754138)<1e-12,
          abs(sum(cme_model$p[cme_model$y==3.875])-0.500363)<1e-12)
cme_model |>
  transmute(`July midpoint` = x, `September midpoint` = y,
            Probability = number(p, accuracy = 0.0001)) |>
  knitr::kable(align = c("r", "r", "r"))
```

The model has correlation `r number(cor_model, accuracy = 0.01)` and covariance `r number(cov_model, accuracy = 0.0001)` squared percentage points.

## Obtain covariance from correlation and spread

Two random variables have standard deviations 2 and 3 and correlation $-0.5$.

Calculate their covariance. What does its sign tell you?

Would correlation alone determine covariance if the standard deviations were unknown?

## Covariance connects to combinations

$$
\operatorname{Cov}(X,Y)=(-0.5)(2)(3)=-3.
$$

The variables have negative linear association; their scales determine the magnitude of covariance.

Independent variables with finite, positive variances have zero covariance and correlation. The converse fails.

Covariance enters the variance of a combination even when the component variables have a nonlinear relationship.
