---
title: "Combining Random Variables"
subtitle: "TEMBA Statistics 2026"
format:
  revealjs:
    chalkboard: true
    embed-resources: false
execute:
  freeze: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

## A portfolio combines two returns

A 60/40 portfolio has 60% of its capital in the first asset and 40% in the second.

| Asset | Portfolio weight | This week's return | Contribution |
|---|---:|---:|---:|
| First | 0.60 | 0.02 | 0.012 |
| Second | 0.40 | −0.01 | −0.004 |
| Portfolio | 1.00 | | 0.008 |

The portfolio earns 0.8%. Its return is a weighted average of the two asset returns.

## Linear combinations

A **linear combination** has the form

$$
Z=aX+bY+c.
$$

A portfolio has weights $a=w$, $b=1-w$, and $c=0$.

Profit can be revenue minus variable costs minus a fixed cost. The fixed cost shifts every possible profit by the same amount.

## Means combine with the same weights

For finite expected values,

$$
E[aX+bY+c]=aE[X]+bE[Y]+c.
$$

The rule holds whether or not the variables are independent.

For a portfolio,

$$
E[R_p]=wE[R_1]+(1-w)E[R_2].
$$

Its expected return is the weighted average of the assets' expected returns.

## Variance includes covariance

For finite variances,

$$
\operatorname{Var}(aX+bY+c)
=a^2\operatorname{Var}(X)+b^2\operatorname{Var}(Y)+2ab\operatorname{Cov}(X,Y).
$$

Each variance receives a squared weight. The covariance term accounts for joint movement; the constant does not change spread.

Zero covariance removes the cross term. Independence is sufficient, but is not required. Take the square root to obtain standard deviation.

## Portfolio variance

$$
\begin{aligned}
\operatorname{Var}(R_p)={}&w^2\operatorname{Var}(R_1)\\
&+(1-w)^2\operatorname{Var}(R_2)\\
&+2w(1-w)\operatorname{Cov}(R_1,R_2).
\end{aligned}
$$

For nonnegative weights, higher covariance increases portfolio variability. Standard deviation measures variation in returns, including both gains and losses.

## Two equally variable stocks

Each stock has expected annual return 10% and annual SD 20%. An equal-weight portfolio has expected return 10%.

$$
\operatorname{Var}(R_p)=0.02+0.02\operatorname{Corr}(R_1,R_2).
$$

| Correlation | Portfolio SD |
|---|---:|
| 0 | ? |
| −0.5 | ? |

Calculate the two standard deviations. Does diversification require negative correlation?

## Correlation changes portfolio variability

```{r}
#| label: table-risk-correlation
#| fig-width: 9
#| fig-height: 3.9
cases <- tibble(correlation = c(1, 0.5, 0, -0.5, -1)) |>
  mutate(risk = sqrt(0.02 + 0.02 * correlation))
cases |>
  transmute(Correlation = correlation,
            `Portfolio standard deviation` = percent(risk, accuracy = 0.1)) |>
  knitr::kable(align = c("r", "r"))
```

The individual stocks each have SD 20%. Even positive correlation below 1 reduces the variability of this equal-weight portfolio.

## Portfolio SD as correlation varies

```{r}
#| label: viz-risk-vs-corr
#| fig-width: 9
#| fig-height: 4.2
risk_curve <- tibble(correlation = seq(-1, 1, length.out = 201)) |>
  mutate(risk = sqrt(0.02 + 0.02 * correlation))
ggplot(risk_curve, aes(correlation, risk)) +
  geom_hline(yintercept = 0.20, linetype = "dashed", color = "grey50") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(data = cases, color = "darkorange", size = 2.5) +
  scale_x_continuous(breaks = seq(-1, 1, 0.5)) +
  scale_y_continuous(labels = percent) +
  labs(x = "Correlation between stock returns",
       y = "Portfolio standard deviation")
```

## The same rules apply to sample summaries

For a column $z_i=ax_i+by_i+c$,

$$
\bar z=a\bar x+b\bar y+c,
$$
$$
s_z^2=a^2s_x^2+b^2s_y^2+2ab s_{xy}.
$$

These identities give the historical mean and variance of the combined column. Using them to describe future returns requires an assumption that the past is informative about the future.

```{r}
#| label: setup-three-funds
#| include: false
portfolios <- read_csv("../data/portfolios/portfolios.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date))

# Annualized summaries used throughout the chapter
funds    <- c("SPY", "EFA", "TLT")
ann_mean <- 52 * sapply(portfolios[funds], mean)
ann_sd   <- sqrt(52) * sapply(portfolios[funds], sd)
ann_cov  <- 52 * cov(portfolios[funds])
cor_mat  <- cor(portfolios[funds])

# ColorBrewer Set2, the book's qualitative palette. Kept as a named vector
# rather than scale_*_brewer because individual colors are pulled out below for
# single-asset panels.
asset_cols <- c(SPY = "#66C2A5", EFA = "#FC8D62", TLT = "#8DA0CB")
```

## Annualized sample summaries

We multiply a weekly sample mean by 52 and a weekly sample SD by $\sqrt{52}$.

The mean is an annualized arithmetic mean, not a compounded annual return. The SD scaling assumes uncorrelated weekly returns with a common variance.

The next comparisons use annualized sample summaries on this same scale.

## Choose a fund to add to SPY

The [three-fund data](../notes/appendix_data.html#sec-data-portfolios-2025) contain `r nrow(portfolios)` weekly returns. Each ETF holds a basket of assets. Summaries are annualized.

```{r}
#| label: table-asset-summary-three
#| fig-width: 9
#| fig-height: 3.9
tibble(Fund = funds,
       Holds = c("U.S. stocks (S&P 500)",
                 "Developed-market stocks",
                 "Long-term U.S. Treasuries"),
       `Mean` = percent(ann_mean, accuracy = 0.1),
       `SD` = percent(ann_sd, accuracy = 0.1)) |>
  knitr::kable(align = c("l", "l", "r", "r"))
```

You currently hold only SPY. Would you add EFA or TLT? Make a provisional choice and state what information you still need.

## How do the funds move with SPY?

```{r}
#| label: viz-scatter-spy-three
#| fig-width: 10
#| fig-height: 4.1
lims <- range(c(portfolios$SPY, portfolios$EFA, portfolios$TLT))

p_efa <- ggplot(portfolios, aes(SPY, EFA)) +
  geom_hline(yintercept = 0, color = "grey85") +
  geom_vline(xintercept = 0, color = "grey85") +
  geom_point(color = asset_cols["EFA"], alpha = 0.7) +
  scale_x_continuous(labels = percent, limits = lims) +
  scale_y_continuous(labels = percent, limits = lims) +
  labs(x = "SPY weekly return", y = "EFA weekly return",
       subtitle = paste0("Corr = ", number(cor_mat["SPY", "EFA"], accuracy = 0.01)))

p_tlt <- ggplot(portfolios, aes(SPY, TLT)) +
  geom_hline(yintercept = 0, color = "grey85") +
  geom_vline(xintercept = 0, color = "grey85") +
  geom_point(color = asset_cols["TLT"], alpha = 0.7) +
  scale_x_continuous(labels = percent, limits = lims) +
  scale_y_continuous(labels = percent, limits = lims) +
  labs(x = "SPY weekly return", y = "TLT weekly return",
       subtitle = paste0("Corr = ", number(cor_mat["SPY", "TLT"], accuracy = 0.01)))

p_efa + p_tlt
```

Revise your choice if needed. Refer to the relationship visible in each plot.

```{r}
#| label: setup-weight-curves-three
#| include: false
mix_stats <- function(other, weights = seq(0, 1, by = 0.005)) {
  tibble(w = weights) |>
    mutate(mean = w * ann_mean["SPY"] + (1 - w) * ann_mean[other],
           sd = sqrt(w^2 * ann_sd["SPY"]^2 + (1 - w)^2 * ann_sd[other]^2 +
                       2 * w * (1 - w) * ann_cov["SPY", other]),
           pair = paste0("SPY + ", other))
}
curves <- bind_rows(mix_stats("EFA"), mix_stats("TLT"))

efa_curve <- filter(curves, pair == "SPY + EFA")
tlt_curve <- filter(curves, pair == "SPY + TLT")

# Optimize over the full interval; the display grid only draws the curves.
w_efa_min <- optimize(function(w) mix_stats("EFA", w)$sd, c(0, 1), tol = 1e-12)$minimum
w_tlt_min <- optimize(function(w) mix_stats("TLT", w)$sd, c(0, 1), tol = 1e-12)$minimum
efa_minvar <- mix_stats("EFA", w_efa_min)
tlt_minvar <- mix_stats("TLT", w_tlt_min)

# Match EFA's minimum variance exactly on TLT's higher-return branch.
target_sd <- efa_minvar$sd
stopifnot(ann_mean["SPY"] > ann_mean["TLT"])
w_tlt_matched <- uniroot(function(w) mix_stats("TLT", w)$sd^2 - target_sd^2,
                         c(w_tlt_min, 1), tol = 1e-12)$root
tlt_matched <- mix_stats("TLT", w_tlt_matched)
matched <- bind_rows(efa_minvar, tlt_matched)
stopifnot(abs(tlt_matched$sd^2 - efa_minvar$sd^2) < 1e-10)

```

## TLT can attain lower portfolio risk

```{r}
#| label: viz-weight-curves-three
#| fig-width: 10
#| fig-height: 3.7
pair_cols <- c("SPY + EFA" = "#66C2A5", "SPY + TLT" = "#FC8D62")
curve_labs <- curves |> filter(near(w, 0.3))

p_mean <- ggplot(curves, aes(w, mean, color = pair)) +
  geom_line(linewidth = 1) +
  geom_text(data = curve_labs, aes(label = pair), vjust = -1, size = 3.2) +
  scale_color_manual(values = pair_cols, guide = "none") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(x = "Weight on SPY", y = "Sample mean return (annualized)")

p_sd <- ggplot(curves, aes(w, sd, color = pair)) +
  geom_line(linewidth = 1) +
  geom_text(data = curve_labs |> mutate(vjust = if_else(pair == "SPY + EFA", -1, 2)),
            aes(label = pair, vjust = vjust), size = 3.2) +
  scale_color_manual(values = pair_cols, guide = "none") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent, limits = c(0, NA)) +
  labs(x = "Weight on SPY", y = "Standard deviation (annualized)")

p_mean + p_sd
```

The minimum annualized sample SD is `r percent(efa_minvar$sd, accuracy = 0.1)` with EFA and `r percent(tlt_minvar$sd, accuracy = 0.1)` with TLT.

## Which portfolio offers more return at the same risk?

Take the **minimum risk attainable with SPY + EFA**: annualized SD `r percent(target_sd, accuracy = 0.1)`.

At that same risk, which portfolio offers the higher expected return: SPY + EFA or SPY + TLT?

Use the historical sample means as estimates of expected returns for this comparison.

## Compare returns at the same risk

```{r}
#| label: table-portfolio-compare-three
#| fig-width: 9
#| fig-height: 3.9
matched |>
  transmute(Portfolio = pair,
            `SPY weight` = percent(w, accuracy = 0.1),
            `Sample mean` = percent(mean, accuracy = 0.1),
            `SD` = percent(sd, accuracy = 0.1)) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

At the same risk, SPY + TLT has an annualized sample mean `r number(100 * (tlt_matched$mean - efa_minvar$mean), accuracy = 0.1)` percentage points higher.

## The matched portfolios on both curves

```{r}
#| label: viz-risk-matched-portfolios
#| fig-width: 10
#| fig-height: 4.2
# Matching colors identify the same allocation in both panels.
matched_mean <- p_mean +
  geom_vline(data = matched, aes(xintercept = w, color = pair),
             linetype = "dotted", linewidth = 0.7, show.legend = FALSE) +
  geom_point(data = matched, size = 3.5) +
  geom_label(data = matched,
             aes(label = percent(mean, accuracy = 0.1)),
             nudge_y = -0.017, size = 4, show.legend = FALSE)

matched_sd <- p_sd +
  geom_hline(yintercept = target_sd, linetype = "dashed", color = "grey35") +
  geom_vline(data = matched, aes(xintercept = w, color = pair),
             linetype = "dotted", linewidth = 0.7, show.legend = FALSE) +
  geom_point(data = matched, size = 3.5) +
  annotate("label", x = 0.5, y = 0.06,
           label = paste0("Both SDs = ", percent(target_sd, accuracy = 0.1)),
           size = 4, color = "grey25")

matched_mean + matched_sd
```

The marked weights are `r percent(efa_minvar$w, accuracy = 0.1)` SPY with EFA and `r percent(tlt_matched$w, accuracy = 0.1)` SPY with TLT. Both have the same SD; the TLT mix has the higher sample mean.

```{r}
#| label: setup-spy-vcit
#| include: false
portfolios <- read_csv("../data/portfolios/portfolios.csv", show_col_types = FALSE)
returns <- read_csv("../data/returns/returns.csv", show_col_types = FALSE)
stopifnot(!anyDuplicated(portfolios$date), !anyDuplicated(returns$date))
fund_data <- portfolios |>
  select(date, SPY, rf) |>
  inner_join(select(returns, date, VCIT), by = "date") |>
  arrange(date)
stopifnot(nrow(fund_data) == nrow(portfolios), !anyNA(fund_data))
funds <- c("SPY", "VCIT")
annual_mean <- 52 * colMeans(fund_data[funds])
annual_cov <- 52 * cov(fund_data[funds])
annual_sd <- sqrt(diag(annual_cov))
fund_cor <- cor(fund_data$SPY, fund_data$VCIT)
rf_annual <- 52 * mean(fund_data$rf)

# w is the fraction invested in SPY. The same weights apply each week.
mix_summary <- function(w) {
  tibble(w = w,
         mean = w * annual_mean["SPY"] + (1-w) * annual_mean["VCIT"],
         sd = sqrt(w^2 * annual_cov["SPY", "SPY"] +
                   (1-w)^2 * annual_cov["VCIT", "VCIT"] +
                   2*w*(1-w) * annual_cov["SPY", "VCIT"])) |>
    mutate(sharpe = (mean-rf_annual)/sd)
}
w_min <- optimize(function(w) mix_summary(w)$sd, c(0, 1),
                  tol = 1e-12)$minimum
sharpe_opt <- optimize(function(w) mix_summary(w)$sharpe, c(0, 1),
                       maximum = TRUE, tol = 1e-12)$maximum
# Include the boundaries when selecting the best allocation.
sharpe_candidates <- mix_summary(c(0, sharpe_opt, 1))
w_sharpe <- sharpe_candidates$w[which.max(sharpe_candidates$sharpe)]
minimum <- mix_summary(w_min)
best_sharpe <- mix_summary(w_sharpe)
curve <- mix_summary(seq(0, 1, length.out = 501))
choices <- mix_summary(c(0, w_min, w_sharpe, 1)) |>
  mutate(allocation = c("All VCIT", "Minimum SD",
                        "Maximum sample Sharpe", "All SPY"))
```

## A second bond fund: VCIT

[VCIT holds intermediate-term investment-grade corporate bonds](../notes/appendix_data.html#sec-data-returns-2025).

```{r}
#| label: viz-scatter-spy
#| fig-width: 9
#| fig-height: 3.7
ggplot(fund_data, aes(SPY, VCIT)) +
  geom_hline(yintercept = 0, color = "grey85") +
  geom_vline(xintercept = 0, color = "grey85") +
  geom_point(alpha = 0.7, color = "steelblue") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(x = "SPY weekly return", y = "VCIT weekly return",
       subtitle = paste0("Sample correlation = ", number(fund_cor, accuracy = 0.01)))
```

The funds differ in their individual variability as well as in how they move together.

## Historical return and variability

```{r}
#| label: table-asset-summary
#| fig-width: 9
#| fig-height: 3.9
tibble(Fund = funds,
       `Mean` = percent(annual_mean, accuracy = 0.1),
       `SD` = percent(annual_sd, accuracy = 0.1)) |>
  knitr::kable(align = c("l", "r", "r"))
```

The mean and standard deviation are annualized sample summaries. For each allocation, we apply the same weights every week, before trading costs.

## Minimum variability uses the variance rule

```{r}
#| label: viz-weight-curves
#| fig-width: 10
#| fig-height: 4.0
mean_plot <- ggplot(curve, aes(w, mean)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = w_min, linetype = "dashed", color = "grey50") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(x = "Weight on SPY", y = "Annualized sample mean")
sd_plot <- ggplot(curve, aes(w, sd)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = w_min, linetype = "dashed", color = "grey50") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(x = "Weight on SPY", y = "Annualized sample standard deviation")
mean_plot + sd_plot
```

The lowest sample SD occurs at `r percent(w_min, accuracy = 0.1)` SPY. Choosing this weight does not use either fund's mean return.

## Return above a benchmark

A portfolio has expected annual return 8%, annual SD 12%, and an annual risk-free benchmark of 3%.

How much expected return does it offer above the benchmark, per unit of standard deviation?

If the benchmark rises to 4%, which input to that comparison changes?

## The Sharpe ratio

For a positive SD and a constant risk-free return over the same period,

$$
\text{Sharpe ratio}=\frac{E[R_p]-r_f}{\operatorname{SD}(R_p)}.
$$

The numerator is expected return above the benchmark. The ratio expresses that excess return per unit of variability.

For the example, $(0.08-0.03)/0.12\approx0.417$. With a 4% benchmark it becomes about 0.333.

## Sample Sharpe ratios across allocations

We use sample means and SDs, with the sample-average Treasury-bill rate of `r percent(rf_annual, accuracy = 0.1)` annualized as a constant benchmark.

```{r}
#| label: viz-sharpe-weights
#| fig-width: 9
#| fig-height: 3.8
ggplot(curve, aes(w, sharpe)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(data = best_sharpe, color = "darkorange", size = 3) +
  scale_x_continuous(labels = percent) +
  labs(x = "Weight on SPY", y = "Sample Sharpe ratio")
```

The maximizing allocation holds `r percent(w_sharpe, accuracy = 0.1)` SPY in this sample.

## Two analysts recommend different weights

One analyst recommends `r percent(w_min, accuracy = 0.1)` SPY; another recommends `r percent(w_sharpe, accuracy = 0.1)` SPY.

What objective would you ask each analyst to state?

Would a higher risk-free benchmark change the allocation with minimum SD? Could it change the allocation with maximum Sharpe ratio?

## Two allocation objectives

```{r}
#| label: table-portfolio-compare
#| fig-width: 9
#| fig-height: 3.9
choices |>
  transmute(Allocation = allocation,
            `SPY weight` = percent(w, accuracy = 0.1),
            `Mean` = percent(mean, accuracy = 0.1),
            `SD` = percent(sd, accuracy = 0.1),
            `Sharpe` = number(sharpe, accuracy = 0.01)) |>
  knitr::kable(align = c("l", "r", "r", "r", "r"))
```

Minimum SD uses variances and covariance. Maximum sample Sharpe also uses mean returns and the benchmark.

These weights optimize the same sample used to report performance; they do not establish the best future allocation.

## Profit is a difference

Revenue has SD 200 and variable costs have SD 120, in thousands of dollars. Their correlation is 0.6.

Which expression gives the variance of revenue minus variable costs?

$$
\begin{aligned}
A:&\quad 200^2+120^2+2(0.6)(200)(120)\\
B:&\quad 200^2+120^2-2(0.6)(200)(120).
\end{aligned}
$$

Explain the effect of positive correlation here, then calculate the profit SD.

## Joint movement can offset in a difference

When revenue and costs rise together, part of their movement cancels in profit. The weights are 1 and $-1$, so **B** applies.

$$
\operatorname{Var}(X-Y)=25{,}600,\qquad
\operatorname{SD}(X-Y)=160.
$$

The profit SD is **\$160,000**. Subtracting a fixed cost changes mean profit but leaves this SD unchanged.
