---
title: "Joint, Marginal, and Conditional Distributions"
filters:
  - details
  - custom-numbered-blocks
custom-numbered-blocks:
  appearances:
    concept:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#F4EAF0', '#7C3F58']
    reference:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#EEF1F3', '#5B6770']
    practice:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#E7F4F5', '#147D85']
  classes:
    Exercise:
      appearance: practice
    Definition:
      appearance: concept
    Notation:
      appearance: reference
    TechnicalDetail:
      appearance: reference
      label: "Technical detail"
      collapse: true
---

```{r}
#| label: setup
#| include: false
#| echo: false
library(tidyverse)
library(scales)

# Figures render at their intrinsic size (fig-width x 96px), capped by the
# ~702px content column -- they are never upscaled to fill it. So axis text
# lands at its true point size, and base_size, not fig-width, is what makes it
# legible next to 17px body text. At 13, tick labels are ~14px.
theme_set(theme_minimal(base_size = 13))

lending_subs <- read_csv("data/lending_club/data/df_subs.csv",
                         show_col_types = FALSE) |>
  mutate(default = as.integer(loan_status == "Charged Off"))

# Historical LendingClub counts and the resulting probability model for a
# future loan. All loan calculations below derive from this single crosstab.
loan_counts <- lending_subs |>
  count(purpose, loan_status) |>
  pivot_wider(names_from = loan_status, values_from = n, values_fill = 0) |>
  transmute(purpose,
            repaid = `Fully Paid`,
            default = `Charged Off`,
            Total = repaid + default) |>
  arrange(desc(Total))

loan_n <- sum(loan_counts$Total)

loan_joint <- loan_counts |>
  transmute(purpose,
            repaid = repaid / loan_n,
            default = default / loan_n)

loan_small <- loan_counts |> filter(purpose == "small_business")
loan_debt <- loan_counts |> filter(purpose == "debt_consolidation")
loan_home <- loan_counts |> filter(purpose == "home_improvement")
loan_default_n <- sum(loan_counts$default)

stopifnot(
  isTRUE(all.equal(sum(loan_joint$repaid) + sum(loan_joint$default), 1)),
  isTRUE(all.equal(rowSums(loan_joint[-1]),
                   loan_counts$Total / loan_n)),
  isTRUE(all.equal(sum(loan_joint$default),
                   loan_default_n / loan_n)),
  isTRUE(all.equal(
    loan_small$default / loan_small$Total,
    (loan_small$default / loan_n) / (loan_small$Total / loan_n)
  ))
)


p3 <- function(x) number(x, accuracy = 0.001)
p4 <- function(x) number(x, accuracy = 0.0001)
```

We want to understand the relationship between default risk and loan purpose. That relationship is an important input into the business decision of whether to write a loan and how to set its terms. Which loan purposes have the greatest default risk?

## Cross-tabulations and contingency tables {#sec-crosstabs}

We'll start with counts, because everything else in this chapter is a count divided by some total.

::: {.Definition #defn-cross-tabulation}
## Cross-tabulation

A **cross-tabulation** or **"crosstab"** --- also called a **contingency table**
--- counts how many observations fall into each *combination* of two variables'
categories. It has one row per category of the first variable, one column per
category of the second, and a count in every cell. Its **margins**, the row and
column of totals, are the frequency tables for the individual variables.
:::

Fourteen purposes make for a bulky table. We want every calculation in this section small enough to check by hand, so we'll narrow to the `r comma(nrow(lending_subs))` loans with three of the more interesting purposes: debt consolidation, home improvement, and small business. @tbl-purpose-default-counts shows their crosstab with loan status.

```{r}
#| label: tbl-purpose-default-counts
#| tbl-cap: "Cross-tabulation of loan purpose and final loan status for the three-purpose subset. Each cell counts the loans with that combination of purpose and outcome; the margins recover the one-variable frequency tables."
#| echo: false
purpose_counts <- lending_subs |>
  count(purpose, loan_status) |>
  pivot_wider(names_from = loan_status, values_from = n, values_fill = 0) |>
  mutate(Total = `Fully Paid` + `Charged Off`) |>
  arrange(desc(Total))

purpose_counts |>
  bind_rows(summarise(purpose_counts, purpose = "**Total**",
                      across(where(is.numeric), sum))) |>
  mutate(across(where(is.numeric), comma)) |>
  rename(Purpose = purpose) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

Look at the margins first, the row and column labeled `Total`. The `Total` column on the right is a frequency table of `purpose` --- the same one-variable summary we built earlier, computed on this subset --- and the bottom row is the frequency table of loan status: `r comma(sum(lending_subs$default == 0))` fully paid, `r comma(sum(lending_subs$default))` charged off. The interior cells record how the two variables tend to move **together**.

With two variables there are three different totals we could divide each count by, and each choice answers a different question. Take the cell for small-business loans that charged off. Dividing it by the grand total gives the share of all loans that are small business *and* charged off:

$$
\frac{`r comma(sum(lending_subs$purpose == "small_business" & lending_subs$default == 1))`}{`r comma(nrow(lending_subs))`} = `r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.001)`.
$$

::: {.Definition #defn-joint-proportion}
## Joint proportion

A **joint proportion** divides an interior cell of a crosstab by the grand
total. It is the share of all observations with that combination of
categories. The collection of all the joint proportions is the **joint
distribution** of the two variables.
:::

Dividing the small-business row total --- the two cells in that row added together --- by the grand total gives the share of all loans that are small business:

$$
\frac{`r comma(sum(lending_subs$purpose == "small_business" & lending_subs$default == 0))` + `r comma(sum(lending_subs$purpose == "small_business" & lending_subs$default == 1))`}{`r comma(nrow(lending_subs))`}
= \frac{`r comma(sum(lending_subs$purpose == "small_business"))`}{`r comma(nrow(lending_subs))`}
= `r number(mean(lending_subs$purpose == "small_business"), accuracy = 0.001)`.
$$

::: {.Definition #defn-marginal-proportion}
## Marginal proportion

A **marginal proportion** divides a row or column total by the grand total. It
gives one category's share without separating the observations by the other
variable. Equivalently, we can add the joint proportions across the
corresponding row or column.
:::

For small-business loans, the marginal proportion is the sum of the two joint
proportions in that row:

$$
`r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 0), accuracy = 0.001)`
+ `r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.001)`
= `r number(mean(lending_subs$purpose == "small_business"), accuracy = 0.001)`.
$$

:::: {.Exercise #exercise-hybrid-work-margins}
## Hybrid work: joint and marginal proportions

Recall the [Bloom, Han, and Liang hybrid-work
experiment](prob_02_probability-basics.qmd#exercise-workplace-random-variables)
from Chapter 7. The researchers assigned Trip.com employees either to work in
the office five days a week or to work from home two days a week [@wfh-bloom].
The table records how many employees in each group stayed at or left the firm:

| Assigned schedule | Stayed | Left | Total |
|---|---:|---:|---:|
| Control | 761 | 59 | 820 |
| Hybrid WFH | 754 | 38 | 792 |
| **Total** | **1,515** | **97** | **1,612** |

1. Find the joint proportion of employees who were assigned to Hybrid WFH
   and left the firm.
2. Find the marginal distribution of assigned schedule.
3. Find the marginal distribution of attrition status.

::: {#solution-hybrid-work-margins .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

The Hybrid WFH/Left cell contains 38 of the 1,612 employees, so its joint
proportion is $38/1612=0.0236$.

The row totals give the marginal distribution of assigned schedule:
$820/1612=0.5087$ for Control and $792/1612=0.4913$ for Hybrid WFH.
The column totals give the marginal distribution of attrition status:
$1515/1612=0.9398$ stayed and $97/1612=0.0602$ left. Each distribution divides
by all 1,612 employees and sums to one.

:::
::::

The crosstab has everything we need to answer the opening question: which loan purposes have the greatest default risk? To compute the small-business default rate, the only relevant information is about small-business loans, so we ignore the other rows and divide by the small-business loan total rather than the grand total. From the counts in @tbl-purpose-default-counts, `r comma(sum(lending_subs$purpose == "small_business" & lending_subs$default == 1))` of the `r comma(sum(lending_subs$purpose == "small_business"))` small-business loans charged off:

$$
\frac{`r comma(sum(lending_subs$purpose == "small_business" & lending_subs$default == 1))`}{`r comma(sum(lending_subs$purpose == "small_business"))`} = `r number(mean(lending_subs$default[lending_subs$purpose == "small_business"]), accuracy = 0.001)`.
$$

::: {.Definition #defn-conditional-proportion}
## Conditional proportion

A **conditional proportion** divides a cell by its row or column total rather
than the grand total. It is the share in one category of one variable among
observations in one category of the other. The conditional proportions across
all categories of the variable being summarized form its **conditional
distribution**.

The direction of the conditioning matters. Among small-business loans, the
share that charged off is
`r number(mean(lending_subs$default[lending_subs$purpose == "small_business"]), accuracy = 0.001)`.
Among charged-off loans, the share issued to small businesses is
`r number(mean(lending_subs$purpose[lending_subs$default == 1] == "small_business"), accuracy = 0.001)`.
Both calculations begin with the same cell but divide by different totals.
:::

@tbl-default-rate-by-purpose collects the default rates for all three purposes.

```{r}
#| label: tbl-default-rate-by-purpose
#| tbl-cap: "Default rate by loan purpose: the proportion of loans charged off, given each purpose. Purposes are sorted from highest to lowest default rate."
#| echo: false
lending_subs |>
  group_by(purpose) |>
  summarise(n = n(), default_rate = mean(default)) |>
  arrange(desc(default_rate)) |>
  mutate(n = comma(n),
         default_rate = number(default_rate, accuracy = 0.001)) |>
  rename(Purpose = purpose, Loans = n, `Default rate` = default_rate) |>
  knitr::kable(align = c("l", "r", "r"))
```

:::: {.Exercise #exercise-hybrid-work-attrition}
## Hybrid work: conditional distributions

Return to the [hybrid-work study](#exercise-hybrid-work-margins):

| Assigned schedule | Stayed | Left | Total |
|---|---:|---:|---:|
| Control | 761 | 59 | 820 |
| Hybrid WFH | 754 | 38 | 792 |
| **Total** | **1,515** | **97** | **1,612** |

Choose one employee at random from the study.

1. Find the conditional distribution of attrition status within each assigned
   schedule. Check that each conditional distribution sums to one.
2. Find $\Pr(\text{Hybrid WFH}\mid\text{Left})$.
3. Explain why the answer to part 2 differs from
   $\Pr(\text{Left}\mid\text{Hybrid WFH})$. Name the denominator in each
   calculation.

::: {#solution-hybrid-work-attrition .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

Within the control group, the conditional distribution of attrition status is

$$
\Pr(\text{Stayed}\mid\text{Control})=\frac{761}{820}=0.9280,
\qquad
\Pr(\text{Left}\mid\text{Control})=\frac{59}{820}=0.0720.
$$

Within the Hybrid WFH group, the corresponding distribution is

$$
\Pr(\text{Stayed}\mid\text{Hybrid WFH})=\frac{754}{792}=0.9520,
\qquad
\Pr(\text{Left}\mid\text{Hybrid WFH})=\frac{38}{792}=0.0480.
$$

Each pair sums to one because every employee in that schedule either stayed or
left. Reversing the condition gives

$$
\Pr(\text{Hybrid WFH}\mid\text{Left})=\frac{38}{97}=0.3918.
$$

The two conditional probabilities share the same numerator, 38 employees, but
use different denominators. The first conditions on the 97 employees who left;
the second conditions on the 792 employees assigned to Hybrid WFH.

:::
::::

## From crosstabs to joint distributions {#sec-joint-marginal-conditional-proportions}

Counts depend on the overall sample size. Dividing every count by the grand
total normalizes the table: the interior cells now sum to one. A proportion and
its percentage are the same number ---
`r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.001)`
is `r percent(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.1)`
--- and from here on we'll write everything as proportions.

@tbl-purpose-default-joint shows the result of dividing every count in @tbl-purpose-default-counts by the grand total, `r comma(nrow(lending_subs))`.

```{r}
#| label: tbl-purpose-default-joint
#| tbl-cap: "Joint proportions of purpose and loan status: each count from the crosstab divided by the grand total. All interior cells together sum to one; the margins give each variable's own distribution."
#| echo: false
joint <- lending_subs |>
  count(purpose, loan_status) |>
  mutate(prop = n / sum(n)) |>
  select(-n) |>
  pivot_wider(names_from = loan_status, values_from = prop) |>
  mutate(Total = `Fully Paid` + `Charged Off`) |>
  arrange(desc(Total))

joint |>
  bind_rows(summarise(joint, purpose = "**Total**",
                      across(where(is.numeric), sum))) |>
  mutate(across(where(is.numeric), \(x) number(x, accuracy = 0.001))) |>
  rename(Purpose = purpose) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

The interior cells are the joint proportions we computed from the count table.
The margins are the corresponding one-variable distributions. As we saw above,
the marginal proportion of small-business loans is the sum of the two joint
proportions in that row:

$$
`r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 0), accuracy = 0.001)`
+ `r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.001)`
= `r number(mean(lending_subs$purpose == "small_business"), accuracy = 0.001)`.
$$

The fully-paid and charged-off small-business loans together account for
`r number(mean(lending_subs$purpose == "small_business"), accuracy = 0.001)` of
all the loans in the table.

The normalized table also gives the conditional proportion. Dividing the joint
proportion for "small business and charged off" by the marginal proportion of
small-business loans gives the same default rate we computed from counts.
Because the table is rounded to three decimals, we carry one extra digit in the
calculation:

$$
\frac{`r number(mean(lending_subs$purpose == "small_business" & lending_subs$default == 1), accuracy = 0.0001)`}{`r number(mean(lending_subs$purpose == "small_business"), accuracy = 0.0001)`} = `r number(mean(lending_subs$default[lending_subs$purpose == "small_business"]), accuracy = 0.001)`.
$$

A conditional proportion is therefore the joint proportion divided by the
marginal proportion of the conditioning variable. Dividing count by count or
proportion by proportion gives the same answer because the grand total cancels.

This normalized crosstab is the observed-data version of the joint probability
tables in the next section. The same correspondence held for one-variable
distributions in @sec-prob-distributions: probabilities for future outcomes
follow the same structure as proportions in observed data. Here it extends to
two variables.

## A probability model for future loans {#sec-loan-probability-model}

So far we have read @tbl-purpose-default-joint as a record of what happened to `r comma(loan_n)` historical loans with these three purposes. If those loans are representative of future loans with the same purposes, the table gives estimated probabilities for a future loan drawn from this restricted population.

:::: {.Exercise #exercise-observed-future-share}
## From an observed share to a future probability

Suppose a lender records 40 defaults among 200 completed loans with the same
three purposes used in this chapter. An analyst computes $40/200=0.20$.

1. What does 0.20 describe about these completed loans?
2. Under what assumption could the analyst use 0.20 as an estimated default
   probability for a future loan from this three-purpose population?

::: {#solution-observed-future-share .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

The value 0.20 is the observed proportion of the 200 loans that defaulted.
It describes those completed loans. To use it as an estimated probability for
a future loan, we assume that these historical loans represent the future
three-purpose population, including its mix of purposes and default behavior.
The observed proportion alone does not establish that assumption.

:::
::::

::: {.Definition #defn-joint-distribution}
## Joint probability distribution

For two discrete random variables, a **joint probability distribution**
assigns a probability to every *pair* of outcomes. Laid out as a table --- one
variable in the rows, the other in the columns --- it follows the same rules
as any probability distribution: every cell is non-negative, and all the cells
together sum to one.
:::

::: {.Notation #notation-loan-outcomes}
## Loan purpose and default

Let $P$ code the purpose of a future loan from this restricted population: $P=1$ for debt consolidation, $P=2$ for home improvement, and $P=3$ for small business. The numbers are labels, not quantities; they let us represent a categorical outcome with a numerical random variable. Let $D$ be the loan's eventual default indicator.
:::

We will use the historical relative frequencies as a probability model for that loan: the probability assigned to each purpose--outcome pair is its count divided by `r comma(loan_n)`. For example,

$$
\Pr(\text{small business and default})
= \Pr(P = 3 \text{ and } D = 1)
= \frac{`r comma(loan_small$default)`}{`r comma(loan_n)`}
= `r number(loan_small$default / loan_n, accuracy = 0.001)`.
$$

Dividing each of the six interior counts by `r comma(loan_n)` produces the
complete joint distribution in @tbl-loan-joint.

```{r}
#| label: tbl-loan-joint
#| tbl-cap: "The joint probability distribution of purpose and outcome for a future loan from the three-purpose population, estimated from the historical relative frequencies."
#| echo: false
loan_joint |>
  mutate(purpose = case_when(
           purpose == "debt_consolidation" ~ "Debt consolidation (P = 1)",
           purpose == "home_improvement" ~ "Home improvement (P = 2)",
           purpose == "small_business" ~ "Small business (P = 3)"
         ),
         across(where(is.numeric), \(x) number(x, accuracy = 0.001))) |>
  rename(Purpose = purpose, `Repaid (D = 0)` = repaid,
         `Default (D = 1)` = default) |>
  knitr::kable(align = c("l", "r", "r"))
```

The six cells cover every possible purpose--outcome pair in this model, and their
probabilities add to `r number(sum(loan_joint$repaid) + sum(loan_joint$default), accuracy = 1)`.
This is Rule 2 from @sec-prob-rules: one of these pairs must occur.

Rule 3 lets us add probabilities for mutually exclusive pairs. In this model, a loan can
default as a debt-consolidation, home-improvement, or small-business loan, so

$$
\begin{aligned}
\Pr(D = 1)
&= \Pr(P = 1 \text{ and } D = 1)
 + \Pr(P = 2 \text{ and } D = 1)
 + \Pr(P = 3 \text{ and } D = 1) \\
&= `r number(loan_debt$default / loan_n, accuracy = 0.001)`
 + `r number(loan_home$default / loan_n, accuracy = 0.001)`
 + `r number(loan_small$default / loan_n, accuracy = 0.001)` \\
&= `r number(loan_default_n / loan_n, accuracy = 0.001)`.
\end{aligned}
$$

The probability of default is therefore the sum of the three joint
probabilities in the default column.

::: {.Definition #defn-marginal-distribution}
## Marginal probability distribution

A **marginal probability distribution** gives the probability distribution of
one variable on its own. We obtain it from a joint distribution by summing over
all possible outcomes of the other variable.
:::

Adding the row and column sums to @tbl-loan-joint gives both marginal
distributions.

```{r}
#| label: tbl-loan-joint-margins
#| tbl-cap: "The joint probability distribution with the marginal distributions of purpose and outcome added in the right column and bottom row."
#| echo: false
loan_joint |>
  mutate(Total = repaid + default) |>
  bind_rows(summarise(loan_joint |> mutate(Total = repaid + default),
                      purpose = "**Total**",
                      across(where(is.numeric), sum))) |>
  mutate(purpose = case_when(
           purpose == "debt_consolidation" ~ "Debt consolidation (P = 1)",
           purpose == "home_improvement" ~ "Home improvement (P = 2)",
           purpose == "small_business" ~ "Small business (P = 3)",
           TRUE ~ purpose
         ),
         across(where(is.numeric), \(x) number(x, accuracy = 0.001))) |>
  rename(Purpose = purpose, `Repaid (D = 0)` = repaid,
         `Default (D = 1)` = default) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

These one-variable distributions appear in the margins of the joint table,
which is where the name *marginal* comes from. For example, summing the two
small-business joint probabilities gives its marginal probability:

$$
\Pr(P = 3)
= `r number(loan_small$repaid / loan_n, accuracy = 0.001)`
+ `r number(loan_small$default / loan_n, accuracy = 0.001)`
= `r number(loan_small$Total / loan_n, accuracy = 0.001)`.
$$

The marginal distribution of $D$ answers "what is the default probability for a future loan with one of these three purposes before we know which one?" But it deliberately averages over purposes, and we know from @tbl-default-rate-by-purpose that purpose matters. The sharper question is conditional: *given* that the loan is for a small business, what is the chance it defaults?

### Conditional probability

::: {.Definition #defn-conditional-probability-joint}
## Conditional probability

For events $A$ and $B$ with $\Pr(B) > 0$, the **conditional probability of $A$
given $B$** is

$$
\Pr(A \mid B) = \frac{\Pr(A \text{ and } B)}{\Pr(B)}.
$$

The bar "$\mid$" reads "given that" or "conditional on." To condition on $B$,
keep only the probability that is consistent with $B$ and rescale it so it sums
to one again. That is what dividing the joint probability by the marginal
probability of $B$ does.
:::

The formula is exactly the joint-over-marginal arithmetic we did with proportions in @sec-joint-marginal-conditional-proportions, now stated as a definition.

On the table, conditioning is a two-step procedure: keep the row or column consistent with the condition, then *renormalize* the remaining probabilities by dividing by their total. Conditioning on $P = 3$ keeps the small-business row. Its total is the marginal probability $\Pr(P = 3)$. Dividing each joint probability in that row by $\Pr(P = 3)$ gives the conditional distribution in @tbl-small-business-conditional.

```{r}
#| label: tbl-small-business-conditional
#| tbl-cap: "The conditional distribution of default status for a future small-business loan. Joint and marginal probabilities are rounded to four decimals; conditional probabilities are computed from the unrounded model."
#| echo: false
loan_small_joint <- loan_joint |> filter(purpose == "small_business")
small_business_marginal <- loan_small_joint$repaid + loan_small_joint$default

tibble(
  Outcome = c("$D = 0$", "$D = 1$", "**Total**"),
  Meaning = c("Loan is repaid", "Loan defaults", ""),
  `Joint probability in the $P = 3$ row` = c(
    p4(loan_small_joint$repaid),
    p4(loan_small_joint$default),
    p4(small_business_marginal)
  ),
  `Divide by $\\Pr(P = 3)$` = c(
    paste0(p4(loan_small_joint$repaid), " / ", p4(small_business_marginal)),
    paste0(p4(loan_small_joint$default), " / ", p4(small_business_marginal)),
    paste0(p4(small_business_marginal), " / ", p4(small_business_marginal))
  ),
  `Conditional probability` = c(
    p3(loan_small_joint$repaid / small_business_marginal),
    p3(loan_small_joint$default / small_business_marginal),
    number(1, accuracy = 0.001)
  )
) |>
  knitr::kable(align = c("l", "l", "r", "r", "r"))
```

Together, the probabilities `r p3(loan_small_joint$repaid / small_business_marginal)` and `r p3(loan_small_joint$default / small_business_marginal)` form the conditional distribution of $D$ given $P = 3$. Because this probability model uses the historical relative frequencies, they match the conditional proportions computed from counts in @sec-crosstabs; @sec-prob-distributions rounded the default probability to 0.25 for its introductory example.

Among the three purposes, small-business loans have the largest default
probability: $\Pr(D = 1 \mid P = 3) =
`r number(loan_small$default / loan_small$Total, accuracy = 0.001)`$. This is
the relevant comparison when we ask which loan purpose has the greatest
default risk.

The reverse conditional answers a different question: among the loans that
defaulted, what share were for small businesses?

$$
\Pr(P = 3 \mid D = 1)
= \frac{`r comma(loan_small$default)`}{`r comma(loan_default_n)`}
= `r number(loan_small$default / loan_default_n, accuracy = 0.001)`.
$$

Small-business loans account for only
`r number(loan_small$Total / loan_n, accuracy = 0.001)` of all loans. Even with
the largest default risk, they make up only
`r number(loan_small$default / loan_default_n, accuracy = 0.001)` of the
defaulted loans.

Most defaulted loans are debt-consolidation loans:

$$
\Pr(P = 1 \mid D = 1)
= \frac{`r comma(loan_debt$default)`}{`r comma(loan_default_n)`}
= `r number(loan_debt$default / loan_default_n, accuracy = 0.001)`.
$$

Debt-consolidation loans have a lower default probability,
$\Pr(D = 1 \mid P = 1) =
`r number(loan_debt$default / loan_debt$Total, accuracy = 0.001)`$, but they
account for `r number(loan_debt$Total / loan_n, accuracy = 0.001)` of all
loans. Their large base rate explains why they also account for most defaults.

:::: {.Exercise #exercise-complete-renewal-model}
## Complete a renewal model

A subscription company uses the following joint probability model for a
randomly selected customer next year. One probability is missing.

| Current tier | Renews | Does not renew | Total |
|---|---:|---:|---:|
| Basic | 0.42 | 0.28 | 0.70 |
| Premium | 0.24 | ? | 0.30 |
| **Total** | ? | ? | **1.00** |

1. Find the missing Premium/Does not renew joint probability.
2. Complete the marginal distribution of renewal status.
3. Find the conditional distribution of renewal status for Basic customers.
4. Find the conditional distribution of renewal status for Premium customers.
5. Find $\Pr(\text{Premium}\mid\text{Renews})$.
6. An analyst says 0.24 is the Premium renewal rate. Explain what 0.24
   represents and give the Premium renewal rate.

::: {#solution-complete-renewal-model .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

The Premium row must sum to its marginal probability of 0.30, so the missing
joint probability is

$$
0.30-0.24=0.06.
$$

Adding down the columns gives the marginal renewal distribution:

$$
\Pr(\text{Renews})=0.42+0.24=0.66,
\qquad
\Pr(\text{Does not renew})=0.28+0.06=0.34.
$$

The completed joint probability model is

| Current tier | Renews | Does not renew | Total |
|---|---:|---:|---:|
| Basic | 0.42 | 0.28 | 0.70 |
| Premium | 0.24 | 0.06 | 0.30 |
| **Total** | **0.66** | **0.34** | **1.00** |

For Basic customers, divide the two joint probabilities in the Basic row by
the row's marginal probability, 0.70:

$$
\Pr(\text{Renews}\mid\text{Basic})=\frac{0.42}{0.70}=0.60,
\qquad
\Pr(\text{Does not renew}\mid\text{Basic})=\frac{0.28}{0.70}=0.40.
$$

For Premium customers, divide by 0.30:

$$
\Pr(\text{Renews}\mid\text{Premium})=\frac{0.24}{0.30}=0.80,
\qquad
\Pr(\text{Does not renew}\mid\text{Premium})=\frac{0.06}{0.30}=0.20.
$$

Among renewing customers, the Premium share is

$$
\Pr(\text{Premium}\mid\text{Renews})=\frac{0.24}{0.66}=0.3636.
$$

The value 0.24 is the joint probability that a customer is Premium and renews.
The Premium renewal rate conditions on Premium customers, so it is
$0.24/0.30=0.80$.

:::
::::

::: {.callout-warning}
## Reversing the condition changes the denominator

Conditional probabilities are asymmetric. The two share a numerator --- the
joint probability of $A$ and $B$ --- but divide it by different *base rates*
(marginal probabilities), so one can be tiny while the other is close to one.
:::

Most tall people aren't in the NBA, but most people in the NBA are tall:
$\Pr(\text{NBA} \mid \text{tall})$ is minuscule while
$\Pr(\text{tall} \mid \text{NBA})$ is close to one. Mathematically,

$$
\Pr(\text{NBA} \mid \text{tall}) = \frac{\Pr(\text{NBA and tall})}{\Pr(\text{tall})},
\qquad
\Pr(\text{tall} \mid \text{NBA}) = \frac{\Pr(\text{NBA and tall})}{\Pr(\text{NBA})}.
$$

Adult heights are roughly normal, so about half of adults are taller than
average. The first conditional therefore divides a tiny joint probability by
about 0.5 and stays tiny. The second divides that same joint probability by the
much smaller probability of being in the NBA, producing a conditional
probability near 1.


## Data used in this section {.unnumbered}

- [LendingClub Accepted Loans](appendix_data.qmd#sec-data-lendingclub-kaggle)
- [Bloom, Han & Liang Hybrid WFH Experiment](appendix_data.qmd#sec-data-wfh-bloom)
