---
title: "Probability Trees, Total Probability, and Bayes' Rule"
filters:
  - details
  - custom-numbered-blocks
custom-numbered-blocks:
  appearances:
    concept:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#F4EAF0', '#7C3F58']
    reference:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#EEF1F3', '#5B6770']
    practice:
      container: foldbox
      boxstyle: foldbox.simple
      collapse: false
      colors: ['#E7F4F5', '#147D85']
  classes:
    Exercise:
      appearance: practice
    Definition:
      appearance: concept
    Notation:
      appearance: reference
    TechnicalDetail:
      appearance: reference
      label: "Technical detail"
      collapse: true
---

```{r}
#| label: setup
#| include: false
#| echo: false
library(tidyverse)
library(scales)

theme_set(theme_minimal(base_size = 13))

# CME FedWatch target-range distributions reported for July 9, 2026. These
# values are the exact rows from CME's historical-probability CSV downloads
# for the July 29 and September 16 meetings. Midpoints stand in for ranges.
cme_marginals <- tribble(
  ~meeting,              ~target_range, ~midpoint, ~probability,
  as.Date("2026-07-29"), "3.50--3.75%", 3.625,     0.754138,
  as.Date("2026-07-29"), "3.75--4.00%", 3.875,     0.245862,
  as.Date("2026-09-16"), "3.50--3.75%", 3.625,     0.376530,
  as.Date("2026-09-16"), "3.75--4.00%", 3.875,     0.500363,
  as.Date("2026-09-16"), "4.00--4.25%", 4.125,     0.123107
)

cme_july_marg <- cme_marginals |>
  filter(meeting == as.Date("2026-07-29")) |>
  transmute(july = target_range, july_mid = midpoint, prob = probability)
cme_sept_marg <- cme_marginals |>
  filter(meeting == as.Date("2026-09-16")) |>
  transmute(sept = target_range, sept_mid = midpoint, prob = probability)

p_july_hold <- cme_july_marg$prob[cme_july_marg$july == "3.50--3.75%"]
p_july_hike <- cme_july_marg$prob[cme_july_marg$july == "3.75--4.00%"]
p_sept_low <- cme_sept_marg$prob[cme_sept_marg$sept == "3.50--3.75%"]
p_sept_3875 <- cme_sept_marg$prob[cme_sept_marg$sept == "3.75--4.00%"]
p_sept_high <- cme_sept_marg$prob[cme_sept_marg$sept == "4.00--4.25%"]

# Under CME's traditional conditional method, both meetings support a hold or
# a 25-basis-point hike in this snapshot. The structural zeros and the two
# reported marginal distributions uniquely determine the four path
# probabilities. Tiny differences between corresponding conditional
# probabilities come from rounding in the published histories.
cme_joint <- tribble(
  ~july,          ~sept,          ~july_mid, ~sept_mid, ~prob,
  "3.50--3.75%",  "3.50--3.75%",  3.625,     3.625,     p_sept_low,
  "3.50--3.75%",  "3.75--4.00%",  3.625,     3.875,     p_july_hold - p_sept_low,
  "3.75--4.00%",  "3.75--4.00%",  3.875,     3.875,     p_july_hike - p_sept_high,
  "3.75--4.00%",  "4.00--4.25%",  3.875,     4.125,     p_sept_high
)

cme_july_marg <- cme_joint |>
  summarise(prob = sum(prob), .by = c(july, july_mid))
cme_sept_marg <- cme_joint |>
  summarise(prob = sum(prob), .by = c(sept, sept_mid))

# Named cells and margins of the CME table, for the prose
cme_p <- function(j, s) {
  v <- cme_joint |> filter(july == j, sept == s) |> pull(prob)
  if (length(v) == 0) 0 else v
}
p_hold_hold <- cme_p("3.50--3.75%", "3.50--3.75%")
p_hold_hike <- cme_p("3.50--3.75%", "3.75--4.00%")
p_hike_hold <- cme_p("3.75--4.00%", "3.75--4.00%")
p_hike_hike <- cme_p("3.75--4.00%", "4.00--4.25%")
cond_hold_hold <- p_hold_hold / p_july_hold
cond_hold_hike <- p_hold_hike / p_july_hold
cond_hike_hold <- p_hike_hold / p_july_hike
cond_hike_hike <- p_hike_hike / p_july_hike

stopifnot(
  isTRUE(all.equal(cond_hold_hold + cond_hold_hike, 1)),
  isTRUE(all.equal(cond_hike_hold + cond_hike_hike, 1)),
  isTRUE(all.equal(p_july_hike * cond_hike_hike, p_hike_hike)),
  isTRUE(all.equal(p_july_hold * cond_hold_hike +
                     p_july_hike * cond_hike_hold,
                   p_sept_3875)),
  isTRUE(all.equal(cme_july_marg$prob,
                   cme_marginals |>
                     filter(meeting == as.Date("2026-07-29")) |>
                     pull(probability))),
  isTRUE(all.equal(cme_sept_marg$prob,
                   cme_marginals |>
                     filter(meeting == as.Date("2026-09-16")) |>
                     pull(probability)))
)

plot_rate_branches <- function(root, outcomes, actions, probs,
                               root_heading, outcome_heading) {
  branches <- tibble(
    x = 0, y = 0, xend = 1, yend = c(1, 0, -1),
    outcome = outcomes, action = actions, prob = probs,
    zero = probs == 0
  )

  ggplot(branches) +
    geom_segment(aes(x = x, y = y, xend = xend, yend = yend,
                     color = zero, linetype = zero), linewidth = 0.7) +
    geom_label(aes(x = 1, y = yend, label = outcome, color = zero),
               fill = "white", label.size = 0.25, size = 3.5) +
    geom_label(aes(x = 0.58, y = 0.58 * yend,
                   label = paste0(action, "\n", number(prob, accuracy = 0.0001)),
                   color = zero),
               fill = "white", label.size = 0.25, size = 3.1) +
    annotate("label", x = 0, y = 0, label = root,
             fill = "white", label.size = 0.25, size = 3.5) +
    annotate("text", x = 0, y = 1.45, label = root_heading,
             fontface = "bold", size = 3.8) +
    annotate("text", x = 1, y = 1.45, label = outcome_heading,
             fontface = "bold", size = 3.8) +
    scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "grey60")) +
    scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed")) +
    guides(color = "none", linetype = "none") +
    coord_cartesian(xlim = c(-0.25, 1.25), ylim = c(-1.25, 1.65),
                    clip = "off") +
    theme_void()
}

p4 <- function(x) number(x, accuracy = 0.0001)
```

## The Fed's two meetings {#sec-fed-two-meetings}

CME FedWatch uses futures prices to estimate probabilities for future federal
funds target ranges. In the July 9, 2026 snapshot we use here, its model allows
a hold or a 25-basis-point hike at each of the next two meetings. A hike of
25 basis points raises the target range by 0.25 percentage points.

::: {.TechnicalDetail #technical-fedwatch-two-meetings}
## How FedWatch constructs the meeting probabilities

FedWatch uses monthly federal funds futures prices to infer the expected change
in the effective federal funds rate at each scheduled meeting. It expresses
that expected change in 25-basis-point increments, then splits probability
between the two adjacent increments that bracket it. For example, an expected
15-basis-point increase becomes a 0.4 probability of a hold and a 0.6
probability of a 25-basis-point hike.

The method does not limit the Fed to a single 25-basis-point move. An expected
72.5-basis-point increase would instead become a 0.1 probability of a
50-basis-point hike and a 0.9 probability of a 75-basis-point hike. Expected
decreases produce the corresponding pair of cuts. [CME's methodology](https://www.cmegroup.com/articles/2023/understanding-the-cme-group-fedwatch-tool-methodology.html)
gives the complete construction.
:::

On July 9, the target range was 3.50--3.75%. Futures prices implied a
probability of `r p4(p_july_hike)` that the Fed would hike at its July 29
meeting and a probability of `r p4(p_july_hold)` that it would hold. The
FedWatch model assigns zero probability to a cut in this snapshot. We can arrange the
nearby outcomes in a branching tree:

```{r}
#| label: fig-cme-july-tree
#| fig-cap: "Market-implied probabilities for the July 29 target range, as reported by CME FedWatch on July 9, 2026. The dashed branch shows a nearby move outside the two supported outcomes in this snapshot."
#| fig-width: 7.3
#| fig-height: 3.2
#| echo: false
plot_rate_branches(
  root = "3.50--3.75%\n(current)",
  outcomes = c("3.25--3.50%", "3.50--3.75%", "3.75--4.00%"),
  actions = c("cut", "hold", "hike"),
  probs = c(0, p_july_hold, p_july_hike),
  root_heading = "July 9",
  outcome_heading = "After July 29 meeting"
)
```

If the Fed were to hike rates at the July meeting, a hold in September
would leave the target at 3.75--4.00%, while another hike would raise it to
4.00--4.25%. We reconstruct the conditional branch probabilities from CME's
published target-range probabilities under this model. The probability of a
September hike given a July hike is `r p4(cond_hike_hike)`:

```{r}
#| label: fig-cme-sept-given-hike
#| fig-cap: "September meeting probabilities after a July hike. Here hold and hike describe the September move; the leaves show the resulting target ranges."
#| fig-width: 7.3
#| fig-height: 3.2
#| echo: false
plot_rate_branches(
  root = "3.75--4.00%\n(after July hike)",
  outcomes = c("3.50--3.75%", "3.75--4.00%", "4.00--4.25%"),
  actions = c("cut", "hold", "hike"),
  probs = c(0, cond_hike_hold, cond_hike_hike),
  root_heading = "After July 29 meeting",
  outcome_heading = "After September 16 meeting"
)
```

These are **conditional probabilities** because they answer an "if" question:
**If** the Fed hikes in July, what is the chance of another hike in September?
The action on a September branch is a move relative to the July target; the
leaf at the end of that branch is the resulting September target range.

If instead the Fed holds the target steady in July, the possible September
outcomes start from 3.50--3.75%. The same September move probabilities now
lead to different target ranges:

```{r}
#| label: fig-cme-sept-given-hold
#| fig-cap: "September meeting probabilities after a July hold. CME applies the same meeting-level move distribution after either July branch."
#| fig-width: 7.3
#| fig-height: 3.2
#| echo: false
plot_rate_branches(
  root = "3.50--3.75%\n(after July hold)",
  outcomes = c("3.25--3.50%", "3.50--3.75%", "3.75--4.00%"),
  actions = c("cut", "hold", "hike"),
  probs = c(0, cond_hold_hold, cond_hold_hike),
  root_heading = "After July 29 meeting",
  outcome_heading = "After September 16 meeting"
)
```

A September hold after a July hike also lands at 3.75--4.00%, giving us a
second path to that target range.

We can collect these possibilities into one tree representing the different
paths the target range could take over the two meetings:

```{r}
#| label: fig-cme-tree-joint
#| fig-cap: "The probability tree implied by CME FedWatch's traditional conditional method. The first branches carry July's marginal probabilities; the second branches apply September's meeting-level move probabilities after each July outcome. Dashed branches show nearby moves outside the supported pair."
#| fig-width: 8.5
#| fig-height: 6
#| echo: false

# Marginal July probabilities (the conditioning denominators)
p_july_hold <- cme_july_marg$prob[cme_july_marg$july == "3.50--3.75%"]
p_july_hike <- cme_july_marg$prob[cme_july_marg$july == "3.75--4.00%"]

# Joint P(July = j, Sept = s); 0 outside the modeled paths
joint_p <- function(j, s) {
  v <- cme_joint |> filter(july == j, sept == s) |> pull(prob)
  if (length(v) == 0) 0 else v
}

# Each July state fans out to a cut, hold, or hike in September.
transitions <- tribble(
  ~from_lab,      ~from_y,  ~to_lab,        ~leaf_y,  ~action,
  "3.50--3.75%",   1.5,     "3.25--3.50%",   2.7,     "cut",
  "3.50--3.75%",   1.5,     "3.50--3.75%",   1.8,     "hold",
  "3.50--3.75%",   1.5,     "3.75--4.00%",   0.9,     "hike",
  "3.75--4.00%",  -1.5,     "3.50--3.75%",  -0.9,     "cut",
  "3.75--4.00%",  -1.5,     "3.75--4.00%",  -1.8,     "hold",
  "3.75--4.00%",  -1.5,     "4.00--4.25%",  -2.7,     "hike"
) |>
  mutate(
    denom = if_else(from_lab == "3.50--3.75%", p_july_hold, p_july_hike),
    cond  = map2_dbl(from_lab, to_lab, joint_p) / denom,
    zero  = cond == 0
  )

# Nodes: root, two July states, six September leaves (leaves may be zero-prob)
nodes <- tribble(
  ~x, ~y,    ~lab,                       ~zero,
  0,  0,     "3.50--3.75%\n(current)",   FALSE,
  1,  1.5,   "3.50--3.75%",              FALSE,
  1, -1.5,   "3.75--4.00%",              FALSE
) |>
  bind_rows(transmute(transitions, x = 2, y = leaf_y, lab = to_lab, zero = zero))

# Edges: root -> July, then each July state -> all three September ranges
edges <- bind_rows(
  tibble(x = 0, y = 0, xend = 1, yend = c(1.5, -1.5), zero = FALSE),
  transmute(transitions, x = 1, y = from_y, xend = 2, yend = leaf_y, zero = zero)
)

branch_labels <- bind_rows(
  tibble(x = 0.5, y = c(0.9, -0.9),
         lab = c(paste0("hold\n", number(p_july_hold, accuracy = 0.0001)),
                 paste0("hike\n", number(p_july_hike, accuracy = 0.0001))),
         zero = FALSE),
  transmute(transitions,
            x = 1.62, y = from_y + 0.6 * (leaf_y - from_y),
            lab = if_else(zero, "0",
                          paste0(action, "\n", number(cond, accuracy = 0.0001))),
            zero = zero)
)

ggplot() +
  geom_segment(data = filter(edges, !zero),
               aes(x = x, y = y, xend = xend, yend = yend),
               color = "grey40") +
  geom_segment(data = filter(edges, zero),
               aes(x = x, y = y, xend = xend, yend = yend),
               color = "grey70", linetype = "dashed") +
  geom_label(data = nodes, aes(x = x, y = y, label = lab, color = zero),
             fill = "white", label.size = 0.25, size = 3.3) +
  geom_label(data = branch_labels, aes(x = x, y = y, label = lab, color = zero),
             fill = "white", alpha = 0.9, size = 2.9) +
  annotate("text", x = 0, y = 3.2, label = "Today (July 9)",
           fontface = "bold", size = 3.6) +
  annotate("text", x = 1, y = 3.2, label = "July 29 meeting",
           fontface = "bold", size = 3.6) +
  annotate("text", x = 2, y = 3.2, label = "September 16 meeting",
           fontface = "bold", size = 3.6) +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "grey60")) +
  guides(color = "none") +
  theme_void() +
  xlim(-0.3, 2.3) +
  ylim(-3.0, 3.5)
```

We omitted the July-cut branch from the full tree because a cut was outside the
two moves supported by this snapshot; the dashed September paths remain for
reference. The tree is another way to represent a **joint probability
distribution** over the target ranges after the July and September meetings.
The probabilities on the branches leaving any one node sum to 1.
Its first branches show the marginal distribution for July. The next set of
branches applies the September meeting's move distribution after each July
outcome. Together they let us reconstruct the joint distribution and the
marginal distribution of the September target range.

## Multiplying along a path

Consider the joint probability of two hikes, which corresponds to a
4.00--4.25% target range in September. The probability of a July hike is
`r p4(p_july_hike)`, and *given* a July hike, the probability of another hike
is `r p4(cond_hike_hike)`. We defined conditional probability as

$$
\Pr(A \mid B) = \frac{\Pr(A \text{ and } B)}{\Pr(B)}.
$$

::: {.callout-note #note-product-rule}
## The product rule

For $\Pr(B)>0$, multiplying both sides by $\Pr(B)$ gives the **product rule**:

$$
\Pr(A \text{ and } B) = \Pr(B)\Pr(A \mid B).
$$

A joint probability is the probability of the condition times the probability
of the other event given that condition.
:::

The first factor is the probability of reaching the July branch; the second
is the probability of the September move given that July outcome. Their
product is the probability of the full path. Therefore, the probability of
two consecutive hikes is

$$
\begin{aligned}
&\Pr(\text{July hike and September hike}) \\
&\qquad = \Pr(\text{July hike})
   \Pr(\text{September hike} \mid \text{July hike}) \\
&\qquad = `r p4(p_july_hike)` \times `r p4(cond_hike_hike)`
 \approx `r p4(p_hike_hike)`.
\end{aligned}
$$

:::: {.Exercise #exercise-subscription-path}
## A subscription path

Suppose 30% of visitors to a website start a free trial. Among visitors who
start a trial, 40% buy a subscription. What is the probability that a randomly
selected visitor starts a trial and buys a subscription? Explain why 0.40
alone does not answer this question.

::: {#solution-subscription-path .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

Multiplying along the path gives $0.30(0.40)=0.12$. Thus, 12% of all visitors
start a trial and buy a subscription. The 0.40 conditions on having started a
trial; it describes only the visitors who reach that branch.

:::
::::

## The law of total probability

Now suppose we are interested in the probability that the target range is
3.75--4.00% in September. There are two ways to get there: a hold followed by
a hike, or a hike followed by a hold. The two paths are mutually exclusive, so
we add their joint probabilities:

$$
\begin{aligned}
\Pr(\text{September target } 3.75\text{--}4.00\%)
&= \underbrace{`r p4(p_july_hold)` \times `r p4(cond_hold_hike)`}_{
     \substack{\text{hold in July,}\\\text{then hike in September}}}
 + \underbrace{`r p4(p_july_hike)` \times `r p4(cond_hike_hold)`}_{
     \substack{\text{hike in July,}\\\text{then hold in September}}} \\
&\approx `r p4(p_hold_hike)` + `r p4(p_hike_hold)` \\
&= `r p4(p_sept_3875)`.
\end{aligned}
$$

This calculation recovers CME's reported September probability from the two
paths that lead to that target range.

The tree highlights both complete paths in dark red, with each path's
probability to the right of its September outcome:

```{r}
#| label: fig-cme-total-probability-tree
#| fig-cap: "The two dark-red paths end at a September target range of 3.75--4.00%. Each path probability is the product of its July probability and its conditional September probability. Values are rounded to four decimals; dashed branches have probability zero."
#| fig-width: 11
#| fig-height: 6
#| echo: false

total_paths <- transitions |>
  mutate(selected = to_lab == "3.75--4.00%",
         ink = case_when(selected ~ "darkred", zero ~ "grey65",
                         TRUE ~ "grey40"),
         product = paste0(p4(denom), " × ", p4(cond), " ≈ ",
                          p4(denom * cond)))
total_edges <- bind_rows(
  tibble(x = 0, y = 0, xend = 1, yend = c(1.5, -1.5),
         ink = "darkred", zero = FALSE),
  transmute(total_paths, x = 1, y = from_y, xend = 2,
            yend = leaf_y, ink, zero)
)
total_nodes <- nodes |>
  mutate(ink = if_else(x < 2 | lab == "3.75--4.00%", "darkred",
                       if_else(zero, "grey65", "grey40")))
total_labels <- branch_labels |>
  mutate(ink = c(rep("darkred", 2), total_paths$ink))

ggplot() +
  geom_segment(data = total_edges,
               aes(x, y, xend = xend, yend = yend,
                   color = ink, linetype = zero), linewidth = 0.7) +
  geom_label(data = total_nodes, aes(x, y, label = lab, color = ink),
             fill = "white", size = 3.3) +
  geom_label(data = total_labels, aes(x, y, label = lab, color = ink),
             fill = "white", size = 2.9) +
  geom_text(data = total_paths,
            aes(x = 2.3, y = leaf_y, label = product, color = ink),
            hjust = 0, size = 3.2) +
  annotate("text", x = c(0, 1, 2, 2.95), y = 3.2,
           label = c("Today (July 9)", "July 29 meeting",
                     "September 16 meeting", "Path probability"),
           fontface = "bold", size = 3.5) +
  scale_color_identity() +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed")) +
  guides(linetype = "none") +
  coord_cartesian(xlim = c(-0.3, 3.65), ylim = c(-3, 3.5), clip = "off") +
  theme_void()
```

::: {.callout-note #note-total-probability}
## The law of total probability

The **law of total probability** applies the same calculation to any event
$B$. Suppose $A_1,\ldots,A_k$ are mutually exclusive and exhaustive events:
exactly one of them occurs. For events $A_i$ with positive probability,

$$
\Pr(B)=\sum_{i=1}^k \Pr(A_i)\Pr(B\mid A_i).
$$

Each term is the joint probability of $A_i$ and $B$. Adding these terms accounts
for every way $B$ can occur, just as summing a column of a joint probability
table gives its marginal probability.
:::

Computing the other joint probabilities and summing the rows and columns gives
a complete probability table like the one we started with for LendingClub:

```{r}
#| label: tbl-cme-joint
#| tbl-cap: "The joint distribution implied by CME FedWatch's traditional conditional method for the July and September 2026 meetings, using probabilities reported on July 9, 2026. Values are rounded to four decimals."
#| echo: false
cme_wide <- tibble(
  july = c("3.50--3.75%", "3.75--4.00%"),
  `3.25--3.50%` = c(0, 0),
  `3.50--3.75%` = c(p_hold_hold, 0),
  `3.75--4.00%` = c(p_hold_hike, p_hike_hold),
  `4.00--4.25%` = c(0, p_hike_hike)
) |>
  mutate(`July marginal` = rowSums(across(where(is.numeric))))

cme_wide |>
  bind_rows(summarise(cme_wide, july = "**September marginal**",
                      across(where(is.numeric), sum))) |>
  mutate(across(where(is.numeric), \(x) number(x, accuracy = 0.0001))) |>
  rename(`July \\\\ September` = july) |>
  knitr::kable(align = c("l", "r", "r", "r", "r", "r"))
```

The table and the tree encode the same information about the joint
distribution in different ways. When events unfold sequentially, the tree is
often more convenient because each path records the order in which the events
occur.

The loan table's conditional default distribution changes across purposes. In
the FedWatch construction, the September *move* distribution stays essentially
the same across July branches, even though the resulting September target
range depends on July's target.

:::: {.Exercise #exercise-takeout-total-probability}
## Orders ready on time

Suppose a restaurant receives 70% of its takeout orders through its app and
30% by phone. These are its only two ordering methods. It has 90% of app
orders and 80% of phone orders ready at the promised time.

1. Find the probability of each path that ends with an order ready on time.
2. Find the probability that a randomly selected order is ready on time.
   Explain why you can add the two path probabilities.

::: {#solution-takeout-total-probability .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

The app-and-ready path has probability $0.70(0.90)=0.63$, and the
phone-and-ready path has probability $0.30(0.80)=0.24$. Every order uses
exactly one of these methods, so the paths are mutually exclusive and account
for every order ready on time. Adding gives $0.63+0.24=0.87$: 87% of orders
are ready at the promised time.

:::
::::

## Worked example: Fraud screening {#sec-fraud-screening}

Suppose an online retailer uses a system that flags transactions for manual
review. In our model, 0.5% of transactions are fraudulent. The system flags
95% of fraudulent transactions and 2% of legitimate transactions. Given that
the system flags a transaction, what is the probability that it is fraudulent?

Let $F$ denote the event that a transaction is fraudulent and $A$ the event
that the system flags it. The 95% rate is a share *of fraudulent transactions*,
so it gives $\Pr(A\mid F)$. Our inputs are

$$
\Pr(F)=0.005,\qquad
\Pr(A\mid F)=0.95,\qquad
\Pr(A\mid F^c)=0.02,
$$

where $F^c$ means the transaction is legitimate. The question asks for the
reverse conditional, $\Pr(F\mid A)$.

The tree shows the two possible screening outcomes within each transaction
status and the probability of each complete path:

```{r}
#| label: fig-fraud-screening-tree
#| fig-cap: "Fraud-screening probability tree. Multiplying the branch probabilities gives each joint probability on the right. The two flagged paths contribute to the overall probability of a flag."
#| fig-width: 10
#| fig-height: 4.8
#| echo: false

fraud_paths <- tribble(
  ~status,       ~from_y, ~outcome,      ~leaf_y, ~prior, ~conditional,
  "Fraudulent",   1.1,   "Flagged",       2.0,   0.005,  0.95,
  "Fraudulent",   1.1,   "Not flagged",   0.5,   0.005,  0.05,
  "Legitimate",  -1.1,   "Flagged",      -0.5,   0.995,  0.02,
  "Legitimate",  -1.1,   "Not flagged",  -2.0,   0.995,  0.98
) |>
  mutate(product = paste0(number(prior, accuracy = 0.001), " × ",
                          number(conditional, accuracy = 0.01), " = ",
                          number(prior * conditional, accuracy = 0.00001)))
fraud_edges <- bind_rows(
  tibble(x = 0, y = 0, xend = 1, yend = c(1.1, -1.1)),
  transmute(fraud_paths, x = 1, y = from_y, xend = 2, yend = leaf_y)
)
fraud_nodes <- bind_rows(
  tibble(x = c(0, 1, 1), y = c(0, 1.1, -1.1),
         label = c("Transaction", "Fraudulent", "Legitimate")),
  transmute(fraud_paths, x = 2, y = leaf_y, label = outcome)
)
fraud_labels <- bind_rows(
  tibble(x = 0.48, y = c(0.72, -0.72), label = c("0.005", "0.995")),
  transmute(fraud_paths, x = 1.55, y = from_y + 0.6 * (leaf_y - from_y),
            label = number(conditional, accuracy = 0.01))
)

ggplot() +
  geom_segment(data = fraud_edges,
               aes(x, y, xend = xend, yend = yend), color = "grey40") +
  geom_label(data = fraud_nodes, aes(x, y, label = label),
             fill = "white", size = 3.4) +
  geom_label(data = fraud_labels, aes(x, y, label = label),
             fill = "white", size = 3.1) +
  geom_text(data = fraud_paths, aes(x = 2.3, y = leaf_y, label = product),
            hjust = 0, size = 3.3) +
  annotate("text", x = c(0, 1, 2, 2.95), y = 2.5,
           label = c("Transaction", "Actual status", "Screening outcome",
                     "Path probability"), fontface = "bold", size = 3.5) +
  coord_cartesian(xlim = c(-0.25, 3.65), ylim = c(-2.3, 2.7), clip = "off") +
  theme_void()
```

Fraudulent transactions make up 0.5% of all transactions, and the system flags
95% of that group. Multiplying these two shares gives the joint probability
that a transaction is fraudulent and flagged:

$$
\Pr(F\text{ and }A)=0.005\times0.95=0.00475.
$$

A flag can also come from a legitimate transaction. Adding the probabilities
of these two paths gives the overall probability of a flag:

$$
\begin{aligned}
\Pr(A)
&=\Pr(F)\Pr(A\mid F)+\Pr(F^c)\Pr(A\mid F^c)\\
&=0.005\times0.95+0.995\times0.02\\
&=0.00475+0.01990=0.02465.
\end{aligned}
$$

Among flagged transactions, the fraudulent share is therefore

$$
\Pr(F\mid A)
=\frac{\Pr(F\text{ and }A)}{\Pr(A)}
=\frac{0.00475}{0.02465}
\approx0.193.
$$

About 19.3% of flagged transactions are fraudulent. The system catches 95% of
fraud, but legitimate transactions are much more common, so even their 2%
flag rate produces more flags than the fraudulent transactions do.

We can represent these rates with a hypothetical group of 100,000
transactions. Of the 500 fraudulent transactions, the system flags
$500\times0.95=475$. Of the 99,500 legitimate transactions, it flags
$99{,}500\times0.02=1{,}990$:

| Actual status | Flagged | Not flagged | Total |
|---|---:|---:|---:|
| Fraudulent | 475 | 25 | 500 |
| Legitimate | 1,990 | 97,510 | 99,500 |
| **Total** | **2,465** | **97,535** | **100,000** |

Conditioning on a flag keeps the 2,465 transactions in the Flagged column.
The fraudulent share is $475/2{,}465\approx0.193$. A flag raises the probability
of fraud from 0.005 to about 0.193, even though most flagged transactions are
legitimate.

## Bayes' rule {#sec-bayes-rule}

The fraud calculation starts with a probability of a flag *given fraud* and
uses it to find the probability of fraud *given a flag*. Substituting the
product rule into the conditional-probability definition gives **Bayes' rule**:

$$
\Pr(F\mid A)
=\frac{\Pr(F)\Pr(A\mid F)}{\Pr(A)}.
$$

The numerator is the probability of a fraudulent transaction that produces a
flag. Dividing by the probability of a flag restricts the comparison to the
flagged transactions, as in the conditional tables in
[Chapter 10](prob_05_multivariate-distributions.qmd#defn-conditional-probability-joint).

::: {.callout-note #note-bayes-rule}
## Bayes' rule

Let $A_1,\ldots,A_k$ be mutually exclusive and exhaustive
events with positive probabilities, and suppose $\Pr(B)>0$. Combining the
product rule and the law of total probability gives

$$
\Pr(A_j\mid B)
=\frac{\Pr(A_j)\Pr(B\mid A_j)}
       {\sum_{i=1}^k\Pr(A_i)\Pr(B\mid A_i)}.
$$

We multiply each starting probability $\Pr(A_i)$ by the conditional probability
of the observed event $B$ within that group. These products are the joint
probabilities of each group and $B$. Dividing by their sum makes the
conditional probabilities add to 1, just as we renormalized the selected
entries of a joint table in Chapter 10.
:::

:::: {.Exercise #exercise-email-base-rates}
## The same filter in two inboxes

Suppose an email filter flags 90% of spam messages and 10% of messages that
are not spam. These rates are the same in two inboxes. In Inbox A, 10% of
incoming messages are spam; in Inbox B, 50% are spam.

1. For each inbox, find the probability that a message is spam given that
   the filter flags it.
2. Explain why these probabilities differ even though the filter has the
   same flagging rates in both inboxes.

::: {#solution-email-base-rates .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

For Inbox A, the spam-and-flagged path has probability $0.10(0.90)=0.09$,
and the nonspam-and-flagged path has probability $0.90(0.10)=0.09$. The spam
share among flagged messages is therefore $0.09/(0.09+0.09)=0.50$.

For Inbox B, the corresponding path probabilities are $0.50(0.90)=0.45$ and
$0.50(0.10)=0.05$. The spam share among flagged messages is
$0.45/(0.45+0.05)=0.90$.

Spam is more common among incoming messages in Inbox B, so more of its flags
come from spam. Holding the two conditional flagging rates fixed does not
hold the composition of flagged messages fixed.

:::
::::

:::: {.Exercise #exercise-supplier-risk-tree}
## Supplier risk as a probability tree

A buyer orders 60% of its components from Supplier A and 40% from Supplier B.
The probability that a component is defective is 0.02 given Supplier A and
0.05 given Supplier B.

1. Draw a probability tree whose first branches identify the supplier and
   whose second branches identify whether the component is defective or not
   defective. Label every branch.
2. Compute the probability of each of the four paths.
3. Use the path probabilities to construct the joint probability table of
   supplier and defect status, including both marginal distributions.
4. Find the marginal probability that a component is defective.
5. Given that a component is defective, find the probability that it came from
   Supplier B.
6. Explain why the answer to part 5 exceeds the marginal probability of
   ordering from Supplier B.

::: {#solution-supplier-risk-tree .details .solution-details summary="Solution"
display="show" non-interactive-summary="Solution"}

The first branches have probabilities 0.60 for Supplier A and 0.40 for Supplier
B. The second-stage branch probabilities are 0.02 Defective and 0.98 Not
defective after Supplier A, and 0.05 Defective and 0.95 Not defective after
Supplier B. The completed tree is

```{r}
#| label: fig-supplier-risk-tree
#| fig-cap: "Completed supplier-risk probability tree. The first branches show the marginal supplier probabilities; the second branches show the conditional defect-status probabilities within each supplier."
#| fig-width: 8
#| fig-height: 4.5
#| echo: false

supplier_edges <- tribble(
  ~x, ~y, ~xend, ~yend,
  0,  0, 1,  1.1,
  0,  0, 1, -1.1,
  1,  1.1, 2,  2.0,
  1,  1.1, 2,  0.5,
  1, -1.1, 2, -0.5,
  1, -1.1, 2, -2.0
)

supplier_nodes <- tribble(
  ~x, ~y, ~label,
  0,  0, "Component",
  1,  1.1, "Supplier A",
  1, -1.1, "Supplier B",
  2,  2.0, "Defective",
  2,  0.5, "Not defective",
  2, -0.5, "Defective",
  2, -2.0, "Not defective"
)

supplier_branch_labels <- tribble(
  ~x, ~y, ~label,
  0.48,  0.72, "A\n0.60",
  0.48, -0.72, "B\n0.40",
  1.53,  1.75, "Defective\n0.02",
  1.55,  0.68, "Not defective\n0.98",
  1.55, -0.68, "Defective\n0.05",
  1.53, -1.75, "Not defective\n0.95"
)

ggplot() +
  geom_segment(data = supplier_edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               color = "grey40") +
  geom_label(data = supplier_nodes, aes(x = x, y = y, label = label),
             fill = "white", label.size = 0.25, size = 3.4) +
  geom_label(data = supplier_branch_labels,
             aes(x = x, y = y, label = label),
             fill = "white", label.size = 0, size = 3.0) +
  annotate("text", x = 0, y = 2.45, label = "Random component",
           fontface = "bold", size = 3.6) +
  annotate("text", x = 1, y = 2.45, label = "Supplier",
           fontface = "bold", size = 3.6) +
  annotate("text", x = 2, y = 2.45, label = "Defect status",
           fontface = "bold", size = 3.6) +
  coord_cartesian(xlim = c(-0.25, 2.3), ylim = c(-2.35, 2.6), clip = "off") +
  theme_void()
```

Multiplying along each path gives the four joint probabilities:

$$
\begin{aligned}
\Pr(A\text{ and Defective}) &= 0.60(0.02)=0.012, \\
\Pr(A\text{ and Not defective}) &= 0.60(0.98)=0.588, \\
\Pr(B\text{ and Defective}) &= 0.40(0.05)=0.020, \\
\Pr(B\text{ and Not defective}) &= 0.40(0.95)=0.380.
\end{aligned}
$$

The path probabilities and their margins form the joint table:

| Supplier | Defective | Not defective | Total |
|---|---:|---:|---:|
| A | 0.012 | 0.588 | 0.600 |
| B | 0.020 | 0.380 | 0.400 |
| **Total** | **0.032** | **0.968** | **1.000** |

Adding the two paths that end in a defect gives
$\Pr(\text{Defective})=0.012+0.020=0.032$. Among defective components, the
Supplier B share is

$$
\Pr(B\mid\text{Defective})=\frac{0.020}{0.032}=0.625.
$$

Supplier B's defective path has probability $0.40(0.05)=0.020$, compared with
$0.60(0.02)=0.012$ for Supplier A. Conditioning on a defect therefore raises
Supplier B's share from 40% of all components to 62.5% of defective components.

:::
::::

## Data used in this section {.unnumbered}

- [CME FedWatch historical probabilities](https://www.cmegroup.com/markets/interest-rates/cme-fedwatch-tool.html), July 9, 2026, with the construction documented in [CME's methodology](https://www.cmegroup.com/articles/2023/understanding-the-cme-group-fedwatch-tool-methodology.html)
