---
title: "Visualizing a Single Quantitative Variable"
---

```{r}
#| label: setup
#| include: false
#| echo: false
library(tidyverse)
library(scales)
library(patchwork)
library(ggiraph)

# Figures render at their intrinsic size (fig-width x 96px), capped by the
# ~702px content column -- they are never upscaled to fill it. So axis text
# lands at its true point size, and base_size, not fig-width, is what makes it
# legible next to 17px body text. At 13, tick labels are ~14px.
theme_set(theme_minimal(base_size = 13))

austin_full <- read_csv("data/austin_houses/austin_houses.csv",
                        show_col_types = FALSE)

load1 <- read_csv("data/ercot/ercotproject/Data/load_1.csv",
                  show_col_types = FALSE)
load2 <- read_csv("data/ercot/ercotproject/Data/load_2.csv",
                  show_col_types = FALSE)
load3 <- read_csv("data/ercot/ercotproject/Data/load_3.csv",
                  show_col_types = FALSE)
ercot <- bind_rows(load1, load2, load3) |>
  filter(LOCATION == "North Central (DFW)") |>
  mutate(datetime = parse_date_time(DATEHOUR, orders = c("ymd HM", "ymd HMS")),
         year = year(datetime),
         date = as.Date(datetime)) |>
  filter(year >= 2016, year <= 2020) |>
  summarize(load = mean(LOAD, na.rm = TRUE), .by = date)
rm(load1, load2, load3)

# GMAT Focus Edition total-score percentile concordance, published by GMAC in
# July 2025 (percentiles cover test takers from July 1, 2020 to June 30, 2025).
gmat <- tribble(
  ~score, ~pct,
  805, 100.0, 795, 100.0, 785, 100.0, 775, 100.0, 765, 99.9, 755, 99.9,
  745, 99.7, 735, 99.5, 725, 99.1, 715, 98.8, 705, 98.0, 695, 97.4,
  685, 95.8, 675, 94.8, 665, 92.1, 655, 90.5, 645, 86.7, 635, 81.9,
  625, 79.2, 615, 76.4, 605, 70.3, 595, 67.1, 585, 60.6, 575, 57.4,
  565, 50.9, 555, 47.8, 545, 41.9, 535, 39.2, 525, 34.0, 515, 31.5,
  505, 27.1, 495, 25.1, 485, 21.3, 475, 19.7, 465, 16.6, 455, 15.2,
  445, 12.7, 435, 11.6, 425,  9.6, 415,  8.7, 405,  7.1, 395,  6.4,
  385,  5.2, 375,  4.7, 365,  3.8, 355,  3.4, 345,  2.7, 335,  2.4,
  325,  1.9, 315,  1.7, 305,  1.3, 295,  1.1, 285,  0.8, 275,  0.7,
  265,  0.5, 255,  0.4, 245,  0.3, 235,  0.3, 225,  0.2, 215,  0.2,
  205,  0.1
) |>
  arrange(score) |>
  mutate(quantile = pct / 100)
```

The first thing we want to do with a quantitative variable is see what values it takes and how often. We'll start with a concrete example: sale prices of single-family homes in ZIP code 78752 in Austin, Texas.

```{r}
#| label: austin-zip-subset
austin_zip <- austin_full |>
  filter(zipcode == 78752)
```

We'll look at `r nrow(austin_zip)` home sales from a single ZIP code in Austin. We have a column called `latest_price` recording the most recent sale price in dollars. Our first goal is to *visualize* the distribution of sale prices to get a sense for typical values, extreme values, and other features of the data. The two visualizations we'll use most are the **histogram** and the **density plot**.


## Histograms

A **histogram** lets us visualize the distribution of a variable by dividing the range of the data into bins and counting how many observations fall in each bin. The x-axis shows the price range, the y-axis shows the count, and each bar's height tells you how many homes sold for a price in that bin's range.

```{r}
#| label: fig-hist-price-count
#| fig-cap: "Histogram of sale prices for homes in Austin ZIP code 78752. Hover over a bar to see its price range and count."
n_bins <- 20
breaks <- seq(min(austin_zip$latest_price), max(austin_zip$latest_price),
              length.out = n_bins + 1)
bw <- diff(breaks)[1]

hist_df <- austin_zip |>
  mutate(bin_idx = pmin(findInterval(latest_price, breaks, rightmost.closed = TRUE),
                        n_bins)) |>
  count(bin_idx) |>
  mutate(
    lower  = breaks[bin_idx],
    upper  = breaks[bin_idx + 1],
    mid    = (lower + upper) / 2,
    tip    = paste0(dollar(round(lower), accuracy = 1000), "–",
                   dollar(round(upper), accuracy = 1000), "\n",
                   n, " home", ifelse(n == 1, "", "s"))
  )

p_hist <- ggplot(hist_df, aes(x = mid, y = n, width = bw)) +
  geom_col_interactive(
    aes(tooltip = tip, data_id = bin_idx),
    fill = "steelblue", color = "white"
  ) +
  geom_rug(data = austin_zip, aes(x = latest_price), inherit.aes = FALSE,
           alpha = 0.4) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Count")

girafe(
  ggobj = p_hist,
  options = list(
    opts_hover(css = "fill:darkorange;"),
    opts_tooltip(css = "background:white; padding:6px 8px; border:1px solid #ccc;
                        border-radius:4px; font-size:13px; white-space:pre;")
  )
)
```

Most homes sold for somewhere between \$150,000 and \$600,000, with a cluster around \$250,000--\$450,000. A handful sold for over \$800,000, stretching the right side of the distribution out.

## Density plots {#sec-density-plots}

A **density plot** is basically a smoother version of a histogram. Here's an example, using the same data:
```{r}
#| label: fig-density-price
#| fig-cap: "Density plot of sale prices in ZIP code 78752."
ggplot(austin_zip, aes(x = latest_price)) +
  geom_density(fill = "steelblue", alpha = 0.3) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Density") +
  geom_rug(alpha = 0.4)
```

Instead of a "staircase" of bars, the density plot gives us a smooth curve. The basic idea, however, is the same: the bars in a histogram are high when there are many data points in the range of its x-values, while the density curve is high at prices where there are more observed sales *close to* that price. For example, the density is much higher around \$370,000 than \$180,000 because there are many more homes with sale prices *near* \$370,000 than \$180,000. That is, the data are more *dense* around \$370,000 than around \$180,000 (even though no single house in this dataset sold for \$370,000).

The construction of a density curve is more involved than a histogram's; don't worry about the details.

:::: {.Definition #defn-density-curve}
## Density curve

A **density curve** describes how the observed values of a quantitative
variable are distributed. The total area under the curve is one, and the *area* under the curve over any interval is roughly the proportion of observed values in that interval.
::::

The area property is easiest to see in a big dataset, so let's look at a new example. The `ercot` data records electricity demand --- the **load**, measured in megawatts --- for the North Central region of the Texas power grid, the zone containing Dallas--Fort Worth, over the years 2016 through 2020. ERCOT, the grid's operator, collected the measurements. We average the hourly measurements within each day, giving us `r comma(nrow(ercot))` daily means. @fig-density-area shows the density curve of daily mean load with two regions shaded: each one contains about 10% of the observed days, chosen from different parts of the distribution.

```{r}
#| label: fig-density-area
#| fig-cap: "Density curve of daily mean electricity demand in the DFW region, 2016--2020. Each shaded interval contains about 10% of the observed days. Its area under the density curve is close to 0.10. The upper interval is more than four times wider than the interval near the center."
#| fig-width: 8
#| fig-height: 4.5
load_vals <- ercot$load
dens <- density(load_vals)
dens_df <- tibble(x = dens$x, y = dens$y)

kde_area <- function(lo, hi) {
  seg <- filter(dens_df, x >= lo, x <= hi)
  sum(diff(seg$x) * (head(seg$y, -1) + tail(seg$y, -1)) / 2)
}

q_mid  <- quantile(load_vals, c(0.40, 0.50))
q_tail <- quantile(load_vals, c(0.90, 1.00))
area_mid  <- kde_area(q_mid[1], q_mid[2])
area_tail <- kde_area(q_tail[1], q_tail[2])

ggplot(dens_df, aes(x = x, y = y)) +
  geom_area(data = filter(dens_df, x >= q_mid[1], x <= q_mid[2]),
            fill = "darkorange", alpha = 0.8) +
  geom_area(data = filter(dens_df, x >= q_tail[1], x <= q_tail[2]),
            fill = "darkorange", alpha = 0.8) +
  geom_line(color = "steelblue", linewidth = 1) +
  annotate("text", x = mean(q_mid), y = max(dens_df$y) * 1.08,
           label = paste0("area = ", number(area_mid, accuracy = 0.001),
                          "\n10% of days"),
           size = 3.4, lineheight = 1.1) +
  annotate("segment", x = mean(q_mid), xend = mean(q_mid),
           y = max(dens_df$y) * 1.0, yend = max(dens_df$y) * 0.9) +
  annotate("text", x = mean(q_tail), y = max(dens_df$y) * 0.32,
           label = paste0("area = ", number(area_tail, accuracy = 0.001),
                          "\n10% of days"),
           size = 3.4, lineheight = 1.1) +
  scale_x_continuous(labels = comma_format()) +
  labs(x = "Daily mean load (MW)", y = "Density")
```

The two shaded regions don't look alike. The left one spans only about `r comma(round(diff(q_mid)))` MW, because demand is dense there: a narrow band near the center collects about 10% of the days. The upper region stretches across about `r comma(round(diff(q_tail)))` MW to gather the same share from the thinner right tail. The blue curve continues slightly beyond the largest observed daily mean because the density smoother spreads some mass beyond the observed range. The area-equals-proportion property is only approximate, but it is a useful guide to what the y-axis units mean (such as they are).


### Comparing density plots and histograms

Histograms are, in a sense, closer to the raw data: each bar is a count you could check by hand, which makes them easy to explain and to trust. Density plots are a little more abstract. In exchange, the density curve makes the overall shape easier to see by smoothing what are often small-sample artifacts like low counts and gaps between observed values. Later in this section we'll use the two together.

To see how much the bin choice can matter, here is the same 78752 price data binned three different ways.

```{r}
#| label: fig-binwidth
#| fig-cap: "The same 134 sale prices in ZIP code 78752, binned three ways. (Bin counts include empty intervals, so e.g. the 10-bin histogram will show fewer than 10 bins)"
#| fig-width: 8.5
#| fig-height: 2.8
mk_hist <- function(bins = NULL, binwidth = NULL, boundary = NULL, title = "") {
  ggplot(austin_zip, aes(x = latest_price)) +
    geom_histogram(bins = bins, binwidth = binwidth, boundary = boundary,
                   fill = "steelblue", color = "white") +
    geom_rug(alpha = 0.3) +
    scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(x = NULL, y = NULL, title = title) +
    theme_minimal(base_size = 10)
}

mk_hist(bins = 10, title = "10 bins") |
  mk_hist(bins = 25, title = "25 bins") |
  mk_hist(bins = 60, title = "60 bins")
```

The goal of the histogram is to "smooth out" gaps and other artifacts of smaller sample sizes while providing a faithful representation of the original data. Ten bins look to be too few (although that's somewhat subjective). Sixty bins produces a choppy picture that isn't much better than the rug. The appropriate choice of bins will be context dependent, but any software producing histograms will make use of broadly sensible defaults. It's a good idea to vary the number of bins and see how the picture changes. Typically you'll find there are a broad range of choices that produce similar results visually. A density plot swaps the bin choice for a **bandwidth** --- a smoothing parameter that measures how close two prices are --- and the bandwidth adjusts the curve in much the same way.

```{r}
#| label: fig-density-bw
#| fig-cap: "The same prices as a density curve under three bandwidths."
#| fig-width: 10
#| fig-height: 2.8
mk_dens <- function(adjust, title) {
  ggplot(austin_zip, aes(x = latest_price)) +
    geom_density(adjust = adjust, fill = "steelblue", alpha = 0.3) +
    geom_rug(alpha = 0.3) +
    scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(x = NULL, y = NULL, title = title) +
    theme_minimal(base_size = 10)
}

mk_dens(0.3, "Wiggly (adjust = 0.3)") |
  mk_dens(1,   "Default (adjust = 1)") |
  mk_dens(2.5, "Smooth (adjust = 2.5)")
```

Small bandwidths act like many small bins, and large bandwidths act like a few large bins. The default bandwidth is chosen by some more involved statistical calculations. It's often quite good, particularly in large samples, but again it's a good idea to vary it a little and see how the picture changes. 

The smoothing in the density plot draws less attention to gaps likely due to small sample sizes, e.g. between the two extreme prices and the rest of the data. But by tracking the data a little more closely, the histogram draws attention to extreme values that might be important to the analysis. 

We don't have to choose. If we rescale a histogram to have a total area of one like a density curve, we can plot them on top of one another. In the rescaled histogram, each bar's area --- not its height alone --- represents the proportion of observations in that bin. This also mitigates one other problem with the original histogram: it's difficult to compare across datasets of different sizes. For example, here are two histograms of prices in two different ZIP codes from the Austin housing data:

```{r}
#| label: fig-hist-bigsmall
#| fig-cap: "Count histograms of sale prices in two Austin ZIP codes, drawn with the same $50,000 bin width and count scale."
#| fig-width: 9
#| fig-height: 3.5
austin_full |>
  filter(zipcode %in% c(78748, 78722), latest_price < 1050000) |>
  mutate(zipcode = factor(zipcode,
                          levels = c(78748, 78722),
                          labels = c("78748 (1,100 sales)", "78722 (90 sales)"))) |>
  ggplot(aes(x = latest_price)) +
  geom_histogram(binwidth = 50000, boundary = 0,
                 fill = "steelblue", color = "white") +
  facet_wrap(~zipcode) +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Sale price", y = "Count")
```

Both panels use the same count scale, but comparing their shapes is difficult because 78748 has 1,100 sales while 78722 has only 90. Scaling each histogram to a common area (of one) lets us compare shapes despite the different sample sizes. It also lets us overlay the density curve:


```{r}
#| label: fig-hist-density-overlay
#| fig-cap: "The same two ZIP codes with each histogram rescaled to unit area and its density curve overlaid."
#| fig-width: 9
#| fig-height: 3.5
austin_full |>
  filter(zipcode %in% c(78748, 78722), latest_price < 1050000) |>
  mutate(zipcode = factor(zipcode,
                          levels = c(78748, 78722),
                          labels = c("78748 (1,100 sales)", "78722 (90 sales)"))) |>
  ggplot(aes(x = latest_price)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 50000, boundary = 0,
                 fill = "steelblue", color = "white") +
  geom_density(color = "darkorange", linewidth = 1) +
  facet_wrap(~zipcode) +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(x = "Sale price", y = "Density")
```

Both panels now share a single vertical scale and it's clear that prices in 78722 sit higher and spread wider than in 78748.

## Percentiles, quantiles, and the cumulative distribution function

One more way to visualize a distribution, though we won't use it much: the **(empirical) cumulative distribution function** (CDF). The CDF shows the fraction of observations that fall at or below a given value. For example, suppose 25% of homes sold for \$300,000 or less. At \$300,000, the CDF is 0.25, or 25%.

Before we look more closely at the CDF, let's define a few terms.

::: {.Definition #defn-percentile-quantile}
## Percentile and quantile

The **p-th percentile** of a distribution is a value at or below which p% of
the observations fall (70% of the observations fall at or below the 70th percentile, for example).

A **quantile** is the same value with its rank written
as a proportion instead of a percentage. The 70th percentile and the 0.7
quantile are the same number, as are the 90th percentile and the 0.9 quantile (and so on).
:::

A percentile isn't always uniquely defined, especially in small datasets or datasets where we see the same value many times, but we don't need to worry about that for our purposes. In the hypothetical house-price example above, \$300,000 is the 25th percentile of prices in 78752. A more salient example for many of you will be the percentile ranks reported alongside standardized test scores. For example, if you'd scored 655 on the GMAT Focus Edition in 2025 that would beat about 90% of test takers --- i.e., 655 is the 90th percentile of GMAT Focus scores over the relevant period (2020-2025 [according to GMAC](https://www.gmac.com/-/media/mbasite/gmat-focus-edition/assets-or-collateral/score-concordance-tables.pdf)). If you had scored 625 on the same test, that would beat about 79% of test takers, so 625 is the 79th percentile.

Some percentiles/quantiles are used often enough to get their own names --- the 50th percentile or 0.50 quantile is known as the **median**, for example. You may also see references to *quartiles*, which are the values that split a dataset into four equal parts. The first quartile (Q1) is the 25th percentile (0.25 quantile), the second quartile (Q2) is the median, and the third quartile (Q3) is the 75th percentile (0.75 quantile).

Once you're familiar with them, plots of CDFs can be a nice companion to our existing visualizations because there's no need to choose bins or smoothing parameters at all, and it's easy to read off percentiles (at least approximately) from the plot. Here's the CDF for our Austin 78752 prices:

```{r}
#| label: fig-ecdf-price
#| fig-cap: "Empirical cumulative distribution function of sale prices in Austin ZIP code 78752. Hover to see the percentile and corresponding price."
ecdf_df <- austin_zip |>
  count(latest_price, name = "n_price") |>
  arrange(latest_price) |>
  mutate(
    ecdf = cumsum(n_price) / sum(n_price),
    tip = paste0(
      "Percentile: ", percent(ecdf, accuracy = 0.1), "\n",
      "Price: ", dollar(latest_price)
    )
  )
 p_ecdf <- ggplot(ecdf_df, aes(x = latest_price, y = ecdf)) +
  geom_step(
    aes(group = 1),
    direction = "hv",
    color = "steelblue",
    linewidth = 1
  ) +
  geom_segment_interactive(
    aes(
      x = latest_price, xend = 0,
      y = ecdf, yend = ecdf,
      data_id = latest_price
    ),
    color = "transparent", linewidth = 0.8, linetype = "dashed"
  ) +
  geom_segment_interactive(
    aes(
      x = latest_price, xend = latest_price,
      y = ecdf, yend = 0,
      data_id = latest_price
    ),
    color = "transparent", linewidth = 0.8, linetype = "dashed"
  ) +
  geom_point_interactive(
    aes(tooltip = tip, data_id = latest_price),
    size = 1,
    color = "transparent"
  ) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Sale price", y = "Cumulative percentage")

girafe(
  ggobj = p_ecdf,
  options = list(
    opts_hover(css = "stroke:darkorange; fill:darkorange;"),
    opts_tooltip(css = "background:white; padding:6px 8px; border:1px solid #ccc;
                        border-radius:4px; font-size:13px; white-space:pre;")
  )
)
```

Returning to the GMAT scores, here's their empirical CDF over 2020-2025, [as published by GMAC in 2025](https://www.gmac.com/-/media/mbasite/gmat-focus-edition/assets-or-collateral/score-concordance-tables.pdf):

```{r}
#| label: fig-gmat-ecdf
#| fig-cap: "Cumulative percentile ranks for GMAT Focus Edition total scores, from GMAC's published 2020--2025 concordance. Hover over a point to see the score and its percentile."
gmat_plot <- gmat |>
  mutate(tip = paste0("Score: ", score, "\n",
                      "Percentile: ", pct, "%"))

p_gmat <- ggplot(gmat_plot, aes(x = score, y = quantile)) +
  geom_step(direction = "hv", color = "steelblue", linewidth = 1) +
  geom_point_interactive(
    aes(tooltip = tip, data_id = score),
    size = 1.4, color = "steelblue"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "GMAT Focus Edition total score", y = "Cumulative percentage")

girafe(
  ggobj = p_gmat,
  options = list(
    opts_hover(css = "stroke:darkorange; fill:darkorange;"),
    opts_tooltip(css = "background:white; padding:6px 8px; border:1px solid #ccc;
                        border-radius:4px; font-size:13px; white-space:pre;")
  )
)
```

Run your finger up from a score of 655 on the x-axis to the CDF and you land near 90% on the y-axis: 655 is approximately the 90th percentile.


## Box and violin plots

A **box plot** is a compact representation of a distribution that highlights the three quartiles (25th, 50th, and 75th percentiles) and any extreme values. This is a box plot of the home sales in 78752:
```{r}
#| label: fig-boxplot-price
#| fig-cap: "Box plot of Austin 78752 home prices. Points beyond the whiskers are flagged as potential outliers."
#| fig-height: 5
#| fig.width: 3
#| fig.align: "center"
ggplot(austin_zip, aes(y = latest_price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.3, width = 0.4) +
  scale_y_continuous(labels = dollar_format()) +
  labs(y = "Sale price of homes in 78752") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

The box begins at the 25th percentile and ends at the 75th. The height of the box --- the distance between the 75th and 25th percentiles --- is the **interquartile range** (IQR). The line inside the box marks the median (50th percentile). The whiskers reach out to the smallest and largest values within 1.5 times the IQR of the box edges, and any points past the whiskers are drawn individually as potential **outliers**.



::: {.callout-warning}
## Common mistake: An "outlier" is not (always) an invalid data point

The 1.5-IQR rule is a convention, not a law. A point identified as an outlier
in a box plot might be a perfectly valid observation from a skewed
distribution, or it might be a data-entry error (or something else). You should never delete an outlier just *because* it was flagged by a rule like the 1.5-IQR rule. You should investigate whether it can be corrected and how sensitive your analysis is to the outlying observations.
:::


A **violin plot** is essentially a density plot mirrored on both sides of an axis, often combined with a box plot in the center. Violins are most useful for comparing distributions across groups, which we'll do in @sec-box-violin-by-group. For a single variable, they're a nice way to show both the density shape and the summary statistics.

```{r}
#| label: fig-violin-price
#| fig-cap: "Violin plot of Austin 78752 home prices."
ggplot(austin_zip, aes(x = "", y = latest_price)) +
  geom_violin(fill = "steelblue", alpha = 0.3) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_y_continuous(labels = dollar_format()) +
  labs(x = "", y = "Sale price")
```

## Data used in this section {.unnumbered}

- [Austin House Listings](appendix_data.qmd#sec-data-housing-kaggle)
- [ERCOT Hourly Load and Temperature Panel](appendix_data.qmd#sec-data-ercot-shaffer)
